---
title: "TLS-DSP analysis"
author: "Rosalia Franco Fernandez"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    html_document:
      code_download: true
      dev: "png"
      code_folding: hide
      toc: true
      toc_float:
        collapsed: false
        smooth_scroll: true
      toc_depth: 3
      wrap: sentence
editor_options: 
    markdown: 
    wrap: sentence
---

# Aim
The aim of this script is to analyze data and generate figures for the TLS paper.

# Setup
Load libraries
```{r setup, include=FALSE, message=F, warning=F}
library(NanoStringNCTools)
library(GeomxTools)
library(GeoMxWorkflows)
library(SpatialDecon)
library(GSVA) 
library(msigdbr)
library(ggforce)
library(scales)
library(cowplot) 
library(umap)
library(Rtsne)
library(ggrepel) 
library(scales) 
library(DT)
library(plotly)
library(gridExtra)
library(RColorBrewer)
library(gt)
library(fgsea)
library(pathview)
library(png)
library(viridis)
library(standR)
library(ggalluvial)
library(preprocessCore)
library(Polychrome)
library("ape")
library("ggtree")
library(tidytree)
library(ggsankey)
library(ggpubr)
library(rstatix)
library(tidyverse)
library(grid)
library(psych)
library(corrplot)
library(Cairo)
library(readxl)
library(writexl)
library(ComplexHeatmap)
library(pheatmap)
library(grDevices)
library(DOSE)
library(enrichplot)

options(scipen = 999) 

knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE) 
```

Upload files
```{r select_annotations}
target_Data <- readRDS("cohort1/Analysis/target_Data.RDS")

figdir <- "cohort1/Analysis/Output/"

count_mat <- dplyr::count(pData(target_Data), tissue,patient,region,segment,`slide name`, name, PNAd, `DC-LAMP`, `germinal center`, FDC)

#annotation vectors
ann_c_TLS <- list(region = c("TLS"=brewer.pal(6,"Set1")[3]))
ann_c_TIL <- list(region = c("TIL"=brewer.pal(6,"Set1")[2]))
ann_c_tumor <- list(region = c("Tumor"=brewer.pal(6,"Set1")[1]))
ann_c_TLS_TIL <-  list(region = c("TLS"=brewer.pal(6,"Set1")[3], "TIL"=brewer.pal(6,"Set1")[2]))
ann_c_region <- list(region = c("Tumor"=brewer.pal(6,"Set1")[1], 
                     "TLS"=brewer.pal(6,"Set1")[3], 
                     "TIL"=brewer.pal(6,"Set1")[2]))

yes_no <- c("yes" = "#0F490E", "no" = "#BD2E2E", "NA" = "#FAF9F6")
ann_c_PNAd <- list(PNAd = c(yes_no))
ann_c_DCLAMP <- list("DC-LAMP" = c(yes_no))
ann_c_germc <- list("germinal center" = c(yes_no[1:2]))
ann_c_FDC <- list("FDC" = c(yes_no))
ann_c_BCL6 <- list("BCL6" = c(yes_no))

ann_c_pt <- list(patient = c("P1"=brewer.pal(12,"Set3")[4],
            "P2"=brewer.pal(12,"Set3")[5],	
            "P3"=brewer.pal(12,"Set3")[6],
            "P4"=brewer.pal(12,"Set3")[7],
            "P5"=brewer.pal(12,"Set3")[8],
            "P6"=brewer.pal(12,"Set3")[9],
            "P7"=brewer.pal(12,"Set3")[10],
            "P8"=brewer.pal(12,"Set3")[11],
            "P9"=brewer.pal(12,"Set3")[12],
            "P10"=brewer.pal(12,"Paired")[1]))

ann_c_pt_GC <- list(patient = c("P1"=brewer.pal(12,"Set3")[4],
            "P2"=brewer.pal(12,"Set3")[5],	
            "P3"=brewer.pal(12,"Set3")[6],
            "P4"=brewer.pal(12,"Set3")[7],
            "P5"=brewer.pal(12,"Set3")[8],
            "P6"=brewer.pal(12,"Set3")[9]))


ann_c_tissue <- list(tissue = c("GC"="#F0A0FF", "PM"="#FCD800"))
ann_c_SN <- list('slide name' = c("GC1_2"=brewer.pal(8,"Dark2")[1], 
                                  "GC3_4"=brewer.pal(8,"Dark2")[2], 
                                  "GC5_6"=brewer.pal(8,"Dark2")[3],
                                  "PM1_4"=brewer.pal(8,"Dark2")[4]))

ann_c_maturity <- list("maturity" = c("LA" = "#ABF1FF", "early TLS" = "#4EC1D7","PF-like TLS" = "#14919E","SF-like TLS" = "#0D636C"))

ann_c_maturity2 <- list("maturity2" = c("non-mature" = "#CBB9F9", "mature" = "#9066FA"))
ann_c_mature_pt2 <- list("mature_pt2" = c("non-mature" = "#CBB9F9", "mature" = "#9066FA"))

ann_c_panels = list('panel' = c("Adaptive Immunity"=brewer.pal(8,"Accent")[1], 
                                     "Innate Immunity"=brewer.pal(8,"Accent")[2], 
                                     "Immune Response"=brewer.pal(8,"Accent")[3]))

ann_c_clusters_TLS <- list("clusters" = c("B naive enriched" = "#FF8080", 
                                          "B memory enriched" = "#A0FFA0", 
                                          "CD4 T memory enriched" = "#05BE78"))

ann_c_clusters_TIL <- list("clusters" = c("CD8 T memory enriched" = "#5757F9" , 
                                          "B memory enriched" = "#A0FFA0", 
                                          "macrophage enriched" = "#C0B57B"))

ann_c_clusters_TLS_TIL <- list("clusters" = c("B naive enriched" = "#FF8080", 
                                              "B memory enriched" = "#A0FFA0", 
                                             "CD4 T memory enriched" = "#05BE78", 
                                              "CD8 T memory enriched" = "#5757F9", 
                                              "macrophage enriched" = "#C0B57B"))

ann_c_clusters_regions <- list("clusters" = c("B naive enriched" = "#FF8080", 
                                              "B memory enriched" = "#A0FFA0", 
                                             "CD4 T memory enriched" = "#05BE78", 
                                              "CD8 T memory enriched" = "#5757F9", 
                                              "macrophage enriched" = "#C0B57B",
                                             "NA" = "#FAF9F6"))

theme_default <- function () {
  
    theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(), 
        plot.title = element_text(hjust = 0.5, face = "bold"), 
        legend.title = element_blank()) 
}

theme_umap <- function () {
  
    theme(axis.title = element_text(face = "bold"), 
        plot.title = element_text(hjust = 0.5, face = "bold"),
        aspect.ratio = 1) 
}


theme_bp <- function() {
        theme(legend.title = element_blank(), 
        axis.title.x = element_blank(),
        axis.title.y = element_text(face = "bold", size = 15),
        plot.title=element_text(hjust=0.5, face = "bold"),
        axis.text.x = element_text(vjust = 1, hjust = 1, angle = 45, size = 11),
        axis.text.y = element_text(size = 11)) 
}

save_pheatmap_pdf <- function(x, filename, width=25, height=15) {
     stopifnot(!missing(x))
     stopifnot(!missing(filename))
     pdf(filename, width=width, height=height)
     grid::grid.newpage()
     grid::grid.draw(x$gtable)
     dev.off()
}

```

Define active aois
```{r define_active_aois}
ann <- pData(target_Data) %>% as.data.frame %>% dplyr::select(`slide name`, name:BCL6)

#grab all genes that survived QC
genes <- rownames(exprs(target_Data))

#order annotations
ann_ordered <- ann %>% arrange(region, tissue, patient)

#create aois for all regions
active_aois_TIL <- rownames(ann_ordered)[ann_ordered$region == "TIL"]
active_aois_TLS <- rownames(ann_ordered)[ann_ordered$region == "TLS"]
active_aois_Tumor <- rownames(ann_ordered)[ann_ordered$region == "Tumor"]

active_aois_TLS_GC <- rownames(ann_ordered)[ann_ordered$region == "TLS" & ann_ordered$tissue == "GC"]

#aois based on segment 
active_aois_CD45 <- rownames(ann_ordered)[ann_ordered$segment == "CD45"]
active_aois_PanCK <- rownames(ann_ordered)[ann_ordered$segment == "PanCK"]

rm(ann_ordered)
```

# UMAP
```{r umap}
custom_umap <- umap::umap.defaults
custom_umap$random_state <- 42
# run UMAP

umap_out <-
  umap(t(log2(assayDataElement(target_Data , elt = "q_norm"))),  
       config = custom_umap)
pData(target_Data)[, c("UMAP_1", "UMAP_2")] <- umap_out$layout[, c(1,2)]

ggplot(pData(target_Data),
       aes(x = UMAP_1, y = UMAP_2, color = region, shape = tissue)) +
  geom_point(size = 2) +
  scale_color_manual(values = ann_c_region[[1]]) +
  theme_classic()  +
  theme_umap() 

ggsave("DSP1_umap region_and_tissue.pdf", device = "pdf", height = 4, width = 4, path = figdir)

umap_out <-
  umap(t(log2(assayDataElement(target_Data[,active_aois_TLS], elt = "q_norm"))),  
       config = custom_umap)
pData(target_Data)[active_aois_TLS, c("UMAP_1", "UMAP_2")] <- umap_out$layout[, c(1,2)]

ggplot(pData(target_Data)[active_aois_TLS,],
       aes(x = UMAP_1, y = UMAP_2, color = region, shape = tissue)) +
  geom_point(size = 2) +
  scale_color_manual(values = ann_c_region[[1]]) +
  theme_classic()  +
  theme_umap()

ggsave("DSP1_umap_TLS_only_and_tissue.pdf", device = "pdf", height = 4, width = 4, path = figdir)

```

# Spatial decon
## Preparation
```{r , results='hide', include = F}
#calculate backgrounds 
bg = derive_GeoMx_background(
  norm = assayDataElement(target_Data , elt = "q_norm"),
  probepool = fData(target_Data)$Module,
  negnames = c("NegProbe-CTP01","NegProbe-Kilo","Negative Probe", "NegProbe-WTX" ))
  #negnames = "NegProbe-WTX")

#safeTME
data("safeTME")
data("safeTME.matches")
current_cell_profile<-safeTME
head(current_cell_profile)

#deselect endothelial cells and fibroblasts from matrix
current_cell_profile <- current_cell_profile[,c(-17, -18)]

#assign prettier names
prettynames <- c("macrophages", "mast", "B naive", "B memory", "plasma cells", "CD4 T naive", "CD4 T memory", "CD8 T naive", "CD8 T memory", "NK cells", "pDC", "mDC", "classical monocytes", "non-classical monocytes", "neutrophils", "Treg") 
colnames(current_cell_profile) <- prettynames

#colorcode the cells
cols = c("#A0FFA0", "#FF8080", "#05BE78", "#FFA040", "#5757F9", "#F74ED6", "#F2B77C", "#C0B57B", "#FFFFA0", "#8BABD3",
         "#90BFF9", "#A00000", "#41F0AE", "#EECCF9", "#B856D7", "#F96495")
names(cols) <- str_sort(prettynames)


```

Run spatial decon
```{r}
res <- runspatialdecon(object = target_Data[,active_aois_CD45],
                      norm_elt = "q_norm",                      # normalized data
                      raw_elt = "exprs",                        # expected background counts for every data point in norm
                      #is_pure_tumor = target_Data$istumor,     # identities of the Tumor segments/observations
                      cell_counts = target_Data$nuclei,         # nuclei counts, used to estimate total cells
                      X = current_cell_profile,                  # safeTME matrix, used by default
                      #cellmerges = safeTME.matches,            # safeTME.matches object, used by default
                      #n_tumor_clusters = 5,                    # how many distinct tumor profiles to append to safeTME
)
```

## Abundance/proportions per patient (means)
Prepare dfs for TLS/TIL
```{r dfs with means, results='hide', include = F}
#extract absolute values
absolute <- pmin(t(res$beta))  
absolute %>% as.data.frame %>% arrange(rownames(absolute))

## dfs for proportions
#first make vectors for TLS patient aois
ann2 <- subset(ann, region == "TLS")

#Subset patients
p1  <- rownames(ann2[which(ann2$patient == "P1"),])
p2  <- rownames(ann2[which(ann2$patient == "P2"),])
p3 <- rownames(ann2[which(ann2$patient == "P3"),])
p4 <- rownames(ann2[which(ann2$patient == "P4"),])
p5  <- rownames(ann2[which(ann2$patient == "P5"),])
p6  <- rownames(ann2[which(ann2$patient == "P6"),])
p7  <- rownames(ann2[which(ann2$patient == "P7"),])
p8  <- rownames(ann2[which(ann2$patient == "P8"),])
p9  <- rownames(ann2[which(ann2$patient == "P9"),])
p10  <- rownames(ann2[which(ann2$patient == "P10"),])

#df with absolute values of TLS
absolute_TLS <- data.frame(P1 = rowMeans(absolute[,p1]), P2 = rowMeans(absolute[,p2]), P3 = rowMeans(absolute[,p3]), P4 = rowMeans(absolute[,p4]), P5 = rowMeans(absolute[,p5]), P6 = rowMeans(absolute[,p6]), P7 = rowMeans(absolute[,p7]), P8 = rowMeans(absolute[,p8]), P9 = rowMeans(absolute[,p9]), P10 = rowMeans(absolute[,p10]))

#df for TLS proportions
props_TLS <- as.data.frame(t(absolute_TLS))
props_TLS <- tibble::rownames_to_column(props_TLS, "patient") #put rownames as first column
props_TLS <- cbind(patient = props_TLS[, 1], props_TLS[, -1]/rowSums(props_TLS[, -1])) #calculate proportions
props_TLS_long <- props_TLS %>% tidyr::gather(cells, prop, macrophages:Treg) %>% arrange(patient) #make data suitable for stacked barplot
props_TLS_long$region <- "TLS"
props_TLS_long$code <- paste(props_TLS_long$region, props_TLS_long$patient)

df <- data.frame(mean = apply(props_TLS[,-1], MARGIN = 2, FUN = mean),
                sd = apply(props_TLS[,-1], MARGIN = 2, FUN = sd),
                n = nrow(props_TLS))

write_xlsx(df %>% rownames_to_column, paste0(figdir, "DSP1_props_TLS_mean_sd_per_pt.xlsx"), )


#make vectors for TIL patient aois
ann2 <- subset(ann, region == "TIL")

#Subset patients
p1  <- rownames(ann2[which(ann2$patient == "P1"),])
p2  <- rownames(ann2[which(ann2$patient == "P2"),])
p3 <- rownames(ann2[which(ann2$patient == "P3"),])
p4 <- rownames(ann2[which(ann2$patient == "P4"),])
p5  <- rownames(ann2[which(ann2$patient == "P5"),])
p6  <- rownames(ann2[which(ann2$patient == "P6"),])
p7  <- rownames(ann2[which(ann2$patient == "P7"),])
p8  <- rownames(ann2[which(ann2$patient == "P8"),])
p9  <- rownames(ann2[which(ann2$patient == "P9"),])
p10  <- rownames(ann2[which(ann2$patient == "P10"),])

#df with absolute values of TIL
absolute_TIL <- data.frame(P1 = rowMeans(absolute[,p1]), P2 = rowMeans(absolute[,p2]), P3 = rowMeans(absolute[,p3]), P4 = rowMeans(absolute[,p4]), P5 = rowMeans(absolute[,p5]), P6 = rowMeans(absolute[,p6]), P7 = rowMeans(absolute[,p7]), P8 = absolute[,p8], P9 = rowMeans(absolute[,p9]), P10 = rowMeans(absolute[,p10]))

#df for TIL proportions
props_TIL <- as.data.frame(t(absolute_TIL))
props_TIL <- tibble::rownames_to_column(props_TIL, "patient") #put rownames as first column
props_TIL <- cbind(patient = props_TIL[, 1], props_TIL[, -1]/rowSums(props_TIL[, -1])) #calculate proportions
props_TIL_long <- props_TIL %>% tidyr::gather(cells, prop, macrophages:Treg) %>% arrange(patient) #make data suitable for stacked barplot
props_TIL_long$region <- "TIL"
props_TIL_long$code <- paste(props_TIL_long$region, props_TIL_long$patient)

df <- data.frame(mean = apply(props_TIL[,-1], MARGIN = 2, FUN = mean),
                sd = apply(props_TIL[,-1], MARGIN = 2, FUN = sd),
                n = nrow(props_TIL))

write_xlsx(df %>% rownames_to_column, paste0(figdir, "DSP1_props_TIL_mean_sd_per_pt.xlsx"))

#merge the long props dfs
props_CD45_long <- rbind(props_TLS_long, props_TIL_long)

## Abundance dfs
#make long dfs for abundance
abs_TIL_long <- as.data.frame(t(absolute_TIL))
abs_TIL_long <- tibble::rownames_to_column(abs_TIL_long, "patient") #put rownames as first column
abs_TIL_long <-  tidyr::gather(abs_TIL_long, cells, freq, macrophages:Treg) 
abs_TIL_long$region <- "TIL"
abs_TIL_long$code <- paste(abs_TIL_long$region, abs_TIL_long$patient)

abs_TLS_long <- as.data.frame(t(absolute_TLS))
abs_TLS_long <- tibble::rownames_to_column(abs_TLS_long, "patient") #put rownames as first column
abs_TLS_long <-  tidyr::gather(abs_TLS_long, cells, freq, macrophages:Treg) 
abs_TLS_long$region <- "TLS"
abs_TLS_long$code <- paste(abs_TLS_long$region, abs_TLS_long$patient)

#merged abundance df
abs_CD45_long <- rbind(abs_TLS_long, abs_TIL_long)
```

## Barplots: Proportional per patient TLS/TIL

```{r CD45 proportions}
aspect_ratio <- 0.5
#TLS only
ggplot(props_TLS_long, aes(x = patient, y = prop, fill = cells)) +
  geom_bar(stat = "identity", position = "stack",  width = 0.9)+ 
  scale_fill_manual(values= cols) +
  theme_classic() +
  labs(y = "proportions", x = "samples", title = "TLS")+
  theme_default() +
  annotate("rect", xmin =0.5, xmax = 6.5, ymin = -0.01, ymax = -0.04, fill = "#F0A0FF", color = "grey",) + #gc
  annotate("rect", xmin =6.5, xmax = 10.5, ymin = -0.01, ymax = -0.04, fill = "#FCD800", color = "grey") #pm

ggsave("DSP1_stacked barplot TLS average of patient.pdf", device = "pdf", height = 5, width = 7, path = figdir)

#TIL only
ggplot(props_TIL_long, aes(x = patient, y = prop, fill = cells)) +
  geom_bar(stat = "identity", position = "stack", width = 0.9)+ 
  scale_fill_manual(values= cols) +
  theme_classic() +
  labs(y = "proportions", x = "samples", title = "TIL")+
  theme_default() +
  annotate("rect", xmin =0.5, xmax = 6.5, ymin = -0.01, ymax = -0.04, fill = "#F0A0FF", color = "grey",) + #gc
  annotate("rect", xmin =6.5, xmax = 10.5, ymin = -0.01, ymax = -0.04, fill = "#FCD800", color = "grey")#pm

ggsave("DSP1_stacked barplot TIL average of patient.pdf", device = "pdf", height = 5, width = 7, path = figdir)
```

## Abundance/proportions per roi
make dfs with all data (no means)
```{r df no means, include = F}
absolute <- pmin(t(res$beta))  
absolute %>% as.data.frame %>% arrange(rownames(absolute))

#TLS
ann2 <- subset(ann, region == "TLS")
absolute_TLS_all <- absolute[,rownames(ann2)]
colnames(absolute_TLS_all) <- ann2$name
props_TLS_all <- as.data.frame(t(absolute_TLS_all))
props_TLS_all <- tibble::rownames_to_column(props_TLS_all, "name") #put rownames as first column
props_TLS_all <- cbind(name = props_TLS_all[, 1], props_TLS_all[, -1]/rowSums(props_TLS_all[, -1])) #calculate proportions
props_TLS_all_long <- props_TLS_all %>% tidyr::gather(cells, prop, macrophages:Treg) %>% arrange(name)
rownames(ann2) <- ann2$name
props_TLS_all_long <- left_join(props_TLS_all_long, ann2, by = "name")

df <- data.frame(mean = apply(props_TLS_all[,-1], MARGIN = 2, FUN = mean), 
                 sd = apply(props_TLS_all[,-1], MARGIN = 2, FUN = sd),
                 n = nrow(props_TLS_all))

#saveRDS(props_TLS_all_long, file = "DSP-TLS/Combined/props_TLS_DSP1.RDS")
write_xlsx(df  %>% rownames_to_column, paste0(figdir, "DSP1_props_TLS_mean_sd_per_aoi.xlsx"))

#TIL
ann2 <- subset(ann, region == "TIL")
absolute_TIL_all <- absolute[,rownames(ann2)]
colnames(absolute_TIL_all) <- ann2$name
props_TIL_all <- as.data.frame(t(absolute_TIL_all))
props_TIL_all <- tibble::rownames_to_column(props_TIL_all, "name") #put rownames as first column
props_TIL_all <- cbind(name = props_TIL_all[, 1], props_TIL_all[, -1]/rowSums(props_TIL_all[, -1])) #calculate proportions
props_TIL_all_long <- props_TIL_all %>% tidyr::gather(cells, prop, macrophages:Treg) %>% arrange(name)
rownames(ann2) <- ann2$name
props_TIL_all_long <- left_join(props_TIL_all_long, ann2, by = "name")

df <- data.frame(mean = apply(props_TIL_all[,-1], MARGIN = 2, FUN = mean), 
                 sd = apply(props_TIL_all[,-1], MARGIN = 2, FUN = sd),
                 n = nrow(props_TIL_all))


#saveRDS(props_TIL_all_long, file = "DSP-TLS/Combined/props_TIL_DSP1.RDS")
write_xlsx(df  %>% rownames_to_column, paste0(figdir, "DSP1_props_TIL_mean_sd_per_aoi.xlsx"))
```

## Barplots: proportions spatial decon all rois
TLS
```{r}
#plotting time
ggplot(props_TLS_all_long, aes(x = name, y = prop, fill = cells)) +
  geom_bar(stat = "identity", position = "stack", width = 0.9)+ 
  ggtitle("TLS") +
  scale_fill_manual(values= cols)+
  theme_classic() +
  theme_default()+
  labs(y = "proportions", x = "samples") +
  annotate("rect", xmin =0.5, xmax = 6.5, ymin = -0.01, ymax = -0.03, fill = "#FB8072", color = "grey") + #p1
  annotate("rect", xmin =6.5, xmax = 12.5, ymin = -0.01, ymax = -0.03, fill = "#80B1D3", color = "grey") + #p2
  annotate("rect", xmin =12.5, xmax = 18.5, ymin = -0.01, ymax = -0.03, fill = "#FDB462", color = "grey") + #p3
  annotate("rect", xmin =18.5, xmax = 25.5, ymin = -0.01, ymax = -0.03, fill = "#B3DE69", color = "grey") + #p4
  annotate("rect", xmin =25.5, xmax = 31.5, ymin = -0.01, ymax = -0.03, fill = "#FCCDE5", color = "grey") + #p5
  annotate("rect", xmin =31.5, xmax = 37.5, ymin = -0.01, ymax = -0.03, fill = "#D9D9D9", color = "grey") + #p6
  annotate("rect", xmin =37.5, xmax = 41.5, ymin = -0.01, ymax = -0.03, fill = "#BC80BD", color = "grey") + #p7
  annotate("rect", xmin =41.5, xmax = 45.5, ymin = -0.01, ymax = -0.03, fill = "#CCEBC5", color = "grey") + #p8
  annotate("rect", xmin =45.5, xmax = 48.5, ymin = -0.01, ymax = -0.03, fill = "#FFED6F", color = "grey") + #p9
  annotate("rect", xmin =48.5, xmax = 53.5, ymin = -0.01, ymax = -0.03, fill = "#A6CEE3", color = "grey") + #p10
  annotate("rect", xmin =0.5, xmax = 37.5, ymin = -0.04, ymax = -0.06, fill = "#F0A0FF", color = "grey") + #gc
  annotate("rect", xmin =37.5, xmax = 53.5, ymin = -0.04, ymax = -0.06, fill = "#FCD800", color = "grey") + #pm
  annotate("rect", xmin =0.5, xmax = 53.5, ymin = -0.07, ymax = -0.09, fill = "#4DAF4A", color = "grey") #tls

ggsave("DSP1_stacked barplot TLS individual aois.pdf", device = "pdf", height = 5, width = 7, path = figdir)

```

TIL
```{r}
#plotting time
ggplot(props_TIL_all_long, aes(x = name, y = prop, fill = cells)) +
  geom_bar(stat = "identity", position = "stack", width = 0.9)+ 
  scale_fill_manual(values= cols)+
  ggtitle("TIL") +
  theme_classic() +
  theme_default() +
  labs(y = "proportions", x = "samples") +
  annotate("rect", xmin =0.5, xmax = 2.5, ymin = -0.01, ymax = -0.03, fill = "#FB8072", color = "grey") + #p1
  annotate("rect", xmin =2.5, xmax = 4.5, ymin = -0.01, ymax = -0.03, fill = "#80B1D3", color = "grey") + #p2
  annotate("rect", xmin =4.5, xmax = 6.5, ymin = -0.01, ymax = -0.03, fill = "#FDB462", color = "grey") + #p3
  annotate("rect", xmin =6.5, xmax = 9.5, ymin = -0.01, ymax = -0.03, fill = "#B3DE69", color = "grey") + #p4
  annotate("rect", xmin =9.5, xmax = 11.5, ymin = -0.01, ymax = -0.03, fill = "#FCCDE5", color = "grey") + #p5
  annotate("rect", xmin =11.5, xmax = 13.5, ymin = -0.01, ymax = -0.03, fill = "#D9D9D9", color = "grey") + #p6
  annotate("rect", xmin =13.5, xmax = 15.5, ymin = -0.01, ymax = -0.03, fill = "#BC80BD", color = "grey") + #p7
  annotate("rect", xmin =15.5, xmax = 16.5, ymin = -0.01, ymax = -0.03, fill = "#CCEBC5", color = "grey") + #p8
  annotate("rect", xmin =16.5, xmax = 18.5, ymin = -0.01, ymax = -0.03, fill = "#FFED6F", color = "grey") + #p9
  annotate("rect", xmin =18.5, xmax = 20.5, ymin = -0.01, ymax = -0.03, fill = "#A6CEE3", color = "grey") + #p10
  annotate("rect", xmin =0.5, xmax = 12.5, ymin = -0.04, ymax = -0.06, fill = "#F0A0FF", color = "grey") + #gc
  annotate("rect", xmin =12.5, xmax = 20.5, ymin = -0.04, ymax = -0.06, fill = "#FCD800", color = "grey") + #pm
  annotate("rect", xmin =0.5, xmax = 20.5, ymin = -0.07, ymax = -0.09, fill = "#377EB8", color = "grey") #TIL

ggsave("DSP1_stacked barplot TIL individual aois.pdf", device = "pdf",  height = 5, width = 3.5, path = figdir)

```

# TLS maturity

```{r}
#create a df with custom aoi names
maturity_genes <- c("MS4A1", "CD3E", "FCER2", "CR2", "CXCL13", "BCL6") #"CD180", "GATA2", "EP300","PTCRA"

#aturity_genes <- c("MS4A1", "CD3E", "FCER2", "CR2", "CXCL13", "BCL6", "CR1","CD38", "CD19",  "CXCR5")

#test <- c("HLA-DMA"  ,    "HLA-DMB"   ,   "HLA-DOA"   ,   "HLA-DOB"   ,   "HLA-DPA1"  ,   "HLA-DPB1"   ,  "HLA-DRA"  ) 
#log2(assayDataElement(target_Data[test,active_aois_TLS], elt = "q_norm")) %>% View

#are these in the panel?
in_panel <- subset(genes, genes %in% maturity_genes)
setdiff(maturity_genes, in_panel)

mature_data <- log2(assayDataElement(target_Data[in_panel,active_aois_TLS], elt = "q_norm"))
order <- sort(colnames(mature_data ))
mature_data  <- mature_data [, order] #reorder columns alphabetically
colnames(mature_data ) <- ann$name[which(ann$region == "TLS")]
ann_TLS <- subset(ann, region == "TLS")
rownames(ann_TLS) <- ann_TLS$name

```


```{r fig.width=15, fig.height=10}
#subset for cell whose means > 0.01
means <- colMeans(props_TLS_all[,-1])
means <- names(means[which(means > 0.02)])

df <- props_TLS_all %>% 
  column_to_rownames("name") %>% 
  select(means) %>% t

maturity_matrix <- rbind(mature_data, df)

mature_TLS_hm<- pheatmap(maturity_matrix ,
         scale = "row", # additional normalization by row 
         angle_col = 90,
         main = "TLS maturity", 
         cutree_rows = 3, cutree_cols = 5,
         show_rownames = T, show_colnames = T,
         clustering_method = "complete",
         clustering_distance_cols = "correlation",
         clustering_distance_rows = "correlation",
         breaks = seq(-3, 3, 0.05),
         color = colorRampPalette(c("#003399", "#FFFFE6", "#CC0000"))(120),
         annotation_colors = c(ann_c_TLS, ann_c_pt, ann_c_tissue),
         cellheight = 15, cellwidth = 15,
         labels_col = ann$name[which(ann$region == "TLS")],
         annotation_col = ann_TLS[, c("tissue", "patient", "region")])

clusters <- cutree(mature_TLS_hm$tree_col, k = 5)
table(clusters)
ann_TLS$maturity <- clusters
ann_TLS$maturity[which(ann_TLS$maturity == 1)] <- "SF-like TLS"
ann_TLS$maturity[which(ann_TLS$maturity == 2)] <- "early TLS"
ann_TLS$maturity[which(ann_TLS$maturity == 3)] <- "SF-like TLS"
ann_TLS$maturity[which(ann_TLS$maturity == 4)] <- "PF-like TLS"
ann_TLS$maturity[which(ann_TLS$maturity == 5)] <- "LA"

ann_TLS$maturity2 <- NA
ann_TLS$maturity2[which(ann_TLS$maturity %in% c("LA", "early TLS"))] <- "non-mature"
ann_TLS$maturity2[which(ann_TLS$maturity %in%  c("PF-like TLS", "SF-like TLS"))] <- "mature"

#add maturity to target_Data
pData(target_Data)$maturity <- NA
pData(target_Data)$maturity[which(pData(target_Data)$region == "TLS")] <- ann_TLS$maturity

pData(target_Data)$maturity2 <- NA
pData(target_Data)$maturity2[which(pData(target_Data)$region == "TLS")] <- ann_TLS$maturity2

ann$maturity <- pData(target_Data)$maturity
ann$maturity2 <- pData(target_Data)$maturity2

ann_TLS$IHC_mat <- pData(target_Data)$IHC_is_mature[which(pData(target_Data)$region == "TLS")]

mature_TLS_hm<- pheatmap(maturity_matrix ,
         scale = "row", # additional normalization by row 
         angle_col = 90,
         #main = "TLS maturity", 
         cutree_rows = 3, cutree_cols = 5,
         show_rownames = T, show_colnames = F,
         clustering_method = "complete",
         clustering_distance_cols = "correlation",
         clustering_distance_rows = "correlation",
         breaks = seq(-3, 3, 0.05),
         color = colorRampPalette(c("#003399", "#FFFFE6", "#CC0000"))(120),
         annotation_colors = c(ann_c_TLS, ann_c_pt, ann_c_tissue, ann_c_maturity, ann_c_maturity2),
         cellheight = 15, cellwidth = 15,
         labels_col = ann$name[which(ann$region == "TLS")],
         annotation_col = ann_TLS[, c("maturity2", "maturity", "tissue", "patient", "region")])

save_pheatmap_pdf(x = mature_TLS_hm, file = paste0(figdir,"DSP1_maturity_TLS_hm_final.pdf"))

```
# Scaled heatmaps
```{r }
#TIL
scaled_TIL <- subset(res$beta, rownames(res$beta) %in% active_aois_TIL)
min <- min(scaled_TIL[scaled_TIL > 0])
scaled_TIL <- sweep(scaled_TIL, 1, pmax(apply(scaled_TIL, 1, max), min), "/")
scaled_TIL <- scaled_TIL[,-c(2,10:12, 13:15)] #keep subsets of interest

#assign names to the df
scaled_TIL <- as.data.frame(scaled_TIL) %>% arrange(rownames(scaled_TIL))
rownames(scaled_TIL) <- ann[which(ann$region == "TIL"),]$name

ann_TIL <- subset(ann, region == "TIL")
rownames(ann_TIL) <- ann_TIL$name

#TLS
scaled_TLS <- subset(res$beta, rownames(res$beta) %in% active_aois_TLS)
min <- min(scaled_TLS[scaled_TLS > 0])
scaled_TLS <- sweep(scaled_TLS, 1, pmax(apply(scaled_TLS, 1, max), min), "/")
scaled_TLS <- scaled_TLS[,-c(2, 10:12, 13:15)] #keep subsets of interest

#assign names to the df
scaled_TLS <- as.data.frame(scaled_TLS) %>% arrange(rownames(scaled_TLS))
rownames(scaled_TLS) <- ann[which(ann$region == "TLS"),]$name

```


## Scaled heatmaps of TLS
Scaled heatmap of TLS
```{r fig.height=10, fig.width=15}
scaled_TLS_heatmap <- pheatmap(t(scaled_TLS), 
         main = "Scaled TLS",
        clustering_method = "complete",
        clustering_distance_cols = "correlation",
        clustering_distance_rows = "correlation",
         cutree_cols = 3, cutree_rows = 1,
         fontsize_row = 11,
         show_rownames = T, show_colnames = F,
         angle_col = "90",
         cellheight = 12, cellwidth = 12, 
         border_color = "grey",
         legend_breaks = c(round(seq(0, 1, length.out = 5), 2)[-5], 1),
         legend_labels = c(round(seq(0, 1, length.out = 5), 2)[-5], 1),
         annotation_colors = c(ann_c_TLS, ann_c_pt, ann_c_tissue, ann_c_maturity),
         annotation_col = ann_TLS[, c("maturity", "tissue","patient", "region")])

#define clustermembership scaled TLS
clusters <- cutree(scaled_TLS_heatmap$tree_col, k = 3)
ann_TLS$clusters <- clusters
ann_TLS$clusters[which(ann_TLS$clusters == 1)] <- "B naive enriched"
ann_TLS$clusters[which(ann_TLS$clusters == 2)] <- "B memory enriched"
ann_TLS$clusters[which(ann_TLS$clusters == 3)] <- "CD4 T memory enriched"

#add membership to target_Data
pData(target_Data)$clusters <- NA
pData(target_Data)$clusters[which(pData(target_Data)$region == "TLS")] <- ann_TLS$clusters
ann$clusters <- pData(target_Data)$clusters

save_pheatmap_pdf(x = scaled_TLS_heatmap, file = paste0(figdir,"DSP1_scaled_TLS_cellprops_hm.pdf")) 
```

```{r fig.height=15, fig.width=20}
scaled_TLS_heatmap <- pheatmap(t(scaled_TLS), 
         main = "Scaled TLS",
        clustering_method = "complete",
        clustering_distance_cols = "correlation",
        clustering_distance_rows = "correlation",
         cutree_cols = 3, cutree_rows = 1,
         fontsize_row = 11,
         show_rownames = T, show_colnames = F,
         angle_col = "90",
         cellheight = 12, cellwidth = 12, 
         border_color = "grey",
         legend_breaks = c(round(seq(0, 1, length.out = 5), 2)[-5], 1),
         legend_labels = c(round(seq(0, 1, length.out = 5), 2)[-5], 1),
         annotation_colors = c(ann_c_TLS, ann_c_pt, ann_c_tissue, ann_c_maturity),
         annotation_col = ann_TLS[, c("maturity","tissue", "patient","region")])

save_pheatmap_pdf(x = scaled_TLS_heatmap, file = paste0(figdir,"DSP1_scaled_TLS_cellprops_with cellbased clusters_hm.pdf")) 

```

## Scaled heatmaps of TIL
```{r fig.height=10, fig.width=15}
scaled_TIL_heatmap <- pheatmap(t(scaled_TIL), 
         main = "Scaled TIL",
        clustering_method = "complete",
        clustering_distance_cols = "correlation",
        clustering_distance_rows = "correlation",
         cutree_cols = 3, cutree_rows = 1,
         fontsize_row = 11,
         show_rownames = T, show_colnames = F,
         angle_col = "90",
         cellheight = 12, cellwidth = 12, 
         border_color = "grey",
         legend_breaks = c(round(seq(0, 1, length.out = 5), 2)[-5], 1),
         legend_labels = c(round(seq(0, 1, length.out = 5), 2)[-5], 1),
         annotation_colors = c(ann_c_TIL, ann_c_pt, ann_c_tissue),
         annotation_col = ann_TIL[, c("tissue", "patient","region")])

#define clustermembership scaled TIL
clusters <- cutree(scaled_TIL_heatmap$tree_col, k = 3)
table(clusters)
ann_TIL$clusters <- clusters
ann_TIL$clusters[which(ann_TIL$clusters == 1)] <- "CD8 T memory enriched"
ann_TIL$clusters[which(ann_TIL$clusters == 2)] <- "macrophage enriched"
ann_TIL$clusters[which(ann_TIL$clusters == 3)] <- "B memory enriched"

ann_c_clusters_TIL <- list("clusters" = c("CD8 T memory enriched" = "#5757F9" , "B memory enriched" = "#A0FFA0", "macrophage enriched" = "#C0B57B"))

#add clusters to target_Data
pData(target_Data)$clusters[which(pData(target_Data)$region == "TIL")] <- ann_TIL$clusters
ann$clusters[which(ann$region == "TIL")] <- pData(target_Data)$clusters[which(pData(target_Data)$region == "TIL")]

save_pheatmap_pdf(x = scaled_TIL_heatmap, file = paste0(figdir,"DSP1_scaled_TIL_cellprops_hm.pdf")) 

```

```{r fig.height=10, fig.width=15}
scaled_TIL_heatmap <- pheatmap(t(scaled_TIL), 
         main = "Scaled TIL",
        clustering_method = "complete",
        clustering_distance_cols = "correlation",
        clustering_distance_rows = "correlation",
         cutree_cols = 3, cutree_rows = 1,
         fontsize_row = 11,
         show_rownames = T, show_colnames = T,
         angle_col = "90",
         cellheight = 12, cellwidth = 12, 
         border_color = "grey",
         legend_breaks = c(round(seq(0, 1, length.out = 5), 2)[-5], 1),
         legend_labels = c(round(seq(0, 1, length.out = 5), 2)[-5], 1),
         annotation_colors = c(ann_c_TIL, ann_c_pt, ann_c_tissue, ann_c_clusters_TIL),
         annotation_col = ann_TIL[, c("clusters", "tissue", "patient","region")])

save_pheatmap_pdf(x = scaled_TIL_heatmap, file = paste0(figdir,"DSP1_scaled_TIL_cellprops_with cellbased clusters_hm.pdf")) 
```

# Boxplot subsets per maturity state

```{r fig.width = 10}
my_comparisons <- list(c("LA", "early TLS"), c("LA", "PF-like TLS"), c("LA", "SF-like TLS"),
                        c("early TLS", "PF-like TLS"), c("early TLS", "SF-like TLS"),
                        c("PF-like TLS", "SF-like TLS"))

coi <- c("macrophages", "B naive", "B memory", "plasma cells", "CD4 T naive", "CD4 T memory", "CD8 T naive",
         "CD8 T memory", "classical monocytes", "non-classical monocytes", "Treg")

mat <- ann_TLS %>% select(name, maturity, maturity2)

props_TLS_all_long %<>% left_join(mat, by = "name")

stat_test <- props_TLS_all_long %>%
  filter(cells %in% coi) %>%
  mutate(maturity = factor(maturity, 
                           levels = c("LA", "early TLS", "PF-like TLS", "SF-like TLS"))) %>%
  mutate(cells = factor(cells, 
                        levels = c("B naive", "B memory", "plasma cells", "CD4 T naive", 
                                   "CD4 T memory", "CD8 T naive", "CD8 T memory","Treg","macrophages", 
                                   "classical monocytes", "non-classical monocytes"))) %>%
  group_by(cells) %>%
  pairwise_wilcox_test(prop ~ maturity, 
                       p.adjust.method = "fdr", 
                       comparisons = my_comparisons) %>%
  mutate(p.symbol = p.adj.signif) 

stat_test$p.symbol[stat_test$p.symbol == "****"] <- "***" # maxsignificance set to p < 0.001

props_TLS_all_long %>%  
  filter(cells %in% coi) %>%
  mutate(maturity = factor(maturity, 
                           levels = c("LA", "early TLS", "PF-like TLS", "SF-like TLS"))) %>%
  mutate(cells = factor(cells, 
                        levels = c("B naive", "B memory", "plasma cells", "CD4 T naive", 
                                   "CD4 T memory", "CD8 T naive", "CD8 T memory","Treg","macrophages", 
                                   "classical monocytes", "non-classical monocytes"))) %>%
  ggplot(aes(x = cells, y = prop)) +
    geom_boxplot(aes(fill = maturity),
                 position = position_dodge(0.8), 
                 width = 0.6, 
                 fatten = 1) +
    scale_fill_manual(values = ann_c_maturity[[1]][1:4]) +
    stat_pvalue_manual(stat_test %>% 
                         add_x_position(x = "cells") %>% 
                         add_y_position(step.increase = 0.06), 
                     hide.ns = T,
                     label = "p.symbol", 
                     tip.length = 0.01, label.size = 6) +
    theme_classic() +
    theme_bp()+
    ylab("proportions") +
    scale_x_discrete(labels = label_wrap(10))+
    ylim(-0.05, 1)

table(ann$maturity[which(ann$region == "TLS")])

ggsave("DSP1_boxplots maturity state per pt_all states.pdf", device = "pdf", height = 5, width = 11, path = figdir)

###

stat_test <- props_TLS_all_long %>%
  filter(cells %in% coi) %>%
  mutate(maturity2 = factor(maturity2, 
                           levels = c("non-mature", "mature"))) %>%
  mutate(cells = factor(cells, 
                        levels = c("B naive", "B memory", "plasma cells", "CD4 T naive", 
                                   "CD4 T memory", "CD8 T naive", "CD8 T memory","Treg","macrophages", 
                                   "classical monocytes", "non-classical monocytes"))) %>%
  group_by(cells) %>%
  pairwise_wilcox_test(prop ~ maturity2, 
                       p.adjust.method = "fdr") %>%
  mutate(p.symbol = p.adj.signif) 

stat_test$p.symbol[stat_test$p.symbol == "****"] <- "***" # maxsignificance set to p < 0.001

props_TLS_all_long %>%  
  filter(cells %in% coi) %>%
  mutate(maturity2 = factor(maturity2, 
                           levels = c("non-mature", "mature"))) %>%
  mutate(cells = factor(cells, 
                        levels = c("B naive", "B memory", "plasma cells", "CD4 T naive", 
                                   "CD4 T memory", "CD8 T naive", "CD8 T memory","Treg","macrophages", 
                                   "classical monocytes", "non-classical monocytes"))) %>%
  ggplot(aes(x = cells, y = prop)) +
    geom_boxplot(aes(fill = maturity2),
                 position = position_dodge(0.8), 
                 width = 0.6, 
                 fatten = 1) +
    scale_fill_manual(values = ann_c_maturity2[[1]][1:4]) +
    stat_pvalue_manual(stat_test %>% 
                         add_x_position(x = "cells") %>% 
                         add_y_position(step.increase = 0.06), 
                     hide.ns = T,
                     label = "p.symbol", 
                     tip.length = 0.01, label.size = 6) +
    theme_classic() +
    theme_bp()+
    ylab("proportions") +
    scale_x_discrete(labels = label_wrap(10))+
    ylim(-0.05, 1)

ggsave("DSP1_boxplots maturity state per pt_matVSnotmat.pdf", device = "pdf", height = 5, width = 11, path = figdir)

```


## correlation maturity IHC and DSP

```{r}
df <- as.data.frame(pData(target_Data)) %>% 
  filter(region == "TLS") %>%
  select(name, IHC_maturity, IHC_is_mature, maturity2, maturity) %>%
  #mutate(maturity2 = ifelse(maturity2 =="mature", "yes", "no" )) %>%
  na.omit

table(df$IHC_is_mature, df$maturity2)
table(df$IHC_maturity, df$maturity)
table(df$maturity)

```

# Pathways: TIL vs TLS
## Load files
```{r load data}
pathwayfiles = c("LBL-10043-08_nCounter_PanCancer_Immune_Profiling_Panel_Gene_List.xlsx",
                 "LBL-10244-03_nCounter_Vantage_RNA_Adaptive_Immunity_Panel.xlsx", 
                 "LBL-10248-03_nCounter_Vantage_RNA_Cellular_Profiling_Panel.xlsx",
                 "LBL-10249-03_nCounter_Vantage_RNA_Innate_Immunity_Panel.xlsx",
                 "LBL-10358-06_nCounter_Vantage_3D_RNA_Protein_Immune_Cell_Profiling_Assay_Probe_List_CS.xlsx",
                 "LBL-10809-03_GeoMx_Cancer_Transcriptome_Atlas_Probe_List.xlsx",
                 "LBL-C0269-03_nCounter-Human-Immunology-V2-Panel-Gene-List.xlsx",
                 "Nanostring immune exhaustion gene list.xlsx")

#vector is also order in list
nameextenstions = c("PanCancer_Immune_Profiling_Panel",
                    "RNA_Adaptive_Immunity_Panel",
                    "RNA_Cellular_Profiling_Panel",
                    "RNA_Innate_Immunity_Panel",
                    "RNA_Protein_Immune_Cell_Profiling_Assay",
                    "GeoMx_Cancer_Transcriptome_Atlas",
                    "Human-Immunology-V2-Panel",
                    "Nanostring immune exhaustion gene list")

relevantslides = c("Immune Response Categories",
                   "Annotations",
                   "Annotations",
                   "Annotations",
                   "Immune Response Categories",
                   "Annotations",
                   "Annotations",
                   "Functional Themes")


pathwaylists = list()
for (i in 1:length(pathwayfiles)){
  df = data.frame(read_xlsx(paste("DSP-TLS/Panels/", pathwayfiles[i], sep=''), sheet=relevantslides[i], skip=1, col_types="text"))
  ncol = dim(df)[2] - 2
  pathways = list()
  for (n in 1:ncol){
    Col = df[,n+2]
    genes = Col%in%'+'
    pathways[[n]] = df[genes,1]
  }
  names(pathways) <- paste(names(df)[3:(ncol+2)])
  names(pathways) <- gsub(".", " ", names(df)[3:(ncol+2)], fixed = T)

  pathwaylists[[i]] <- pathways
}

names(pathwaylists) <- nameextenstions

```


``` {r convert to correct format}
genes <- rownames(exprs(target_Data))

#import different categories from CTApanel
categories <- read_xlsx("DSP-TLS/Panels/LBL-10809-03_GeoMx_Cancer_Transcriptome_Atlas_Probe_List.xlsx", sheet = "Categories") %>% as.data.frame
categories <- tidyr::gather(categories, "panel","pathways")
categories <- subset(categories, pathways != "NA")
categories <- arrange(categories, pathways)

#make a list
msigdbr_list <-  pathwaylists[["GeoMx_Cancer_Transcriptome_Atlas"]]

names <- read_xlsx("DSP-TLS/Panels/CTA-panel-names.xlsx") %>% as.data.frame
names(msigdbr_list) <- names$new_names #correct the names

msigdbr_list<- lapply(msigdbr_list, function(x) x[x %in% genes])
msigdbr_list <- msigdbr_list[lapply(msigdbr_list, length) > 3]

```

add extra pathways
```{r}
#Huang X, Nat. Com. 2023
gene_panels <- read_xlsx("General/Suppl_gene signatures.xlsx")
gene_list <- list()

for (sig in 1:ncol(gene_panels)) {
  
  gene_list[[colnames(gene_panels)[sig]]] <- gene_panels[,sig] %>% na.omit %>% unlist %>% as.vector
  
}

gene_list <- lapply(gene_list, function(x) x[x %in% genes])
gene_list <- gene_list [lapply(gene_list, length) > 3]

```

## pathways
Subset immune pathways
```{r}
immune_panels <- subset(msigdbr_list, names(msigdbr_list) %in% categories$pathways[categories$panel == "Adaptive Immunity" | categories$panel == "Innate Immunity" | categories$panel == "Immune Response"])

#add from other CTA panels
other <- subset(msigdbr_list, names(msigdbr_list) %in% categories$pathways[categories$panel == "Cell Function"])
immune_panels["Apoptosis"] <- other["Apoptosis"]
immune_panels["Phagocytosis"] <- other["Phagocytosis"]

other <- subset(msigdbr_list, names(msigdbr_list) %in% categories$pathways[categories$panel == "Signaling Pathways"])
immune_panels["TGF-beta Signaling"] <- other["TGF-beta Signaling"]
immune_panels["VEGF Signaling"] <- other["VEGF Signaling"]

#add from Huang X, Nat. Com. 2023
immune_panels["M1-like activation"] <- gene_list["M1"]
immune_panels["M2-like activation"] <- gene_list["M2"]
immune_panels["Pro-inflammation"] <- gene_list["Pro-inflammatory"]
immune_panels["Anti-inflammation"] <- gene_list["Anti-inflammatory"]
immune_panels["Cytotoxicity"] <- gene_list["Cytotoxic"] #has more genes
immune_panels["Proliferation"] <- gene_list["Proliferative"]

immune_panels["Cancer Antigens"] <- NULL

pw_gene_df<-data.frame(Pathway = names(immune_panels))
pw_gene_df$genes<-immune_panels
```

Run GSVSA
```{r run_gsva, warning = F}
#calculate enrichment of panels
ssgsea_results <- GSVA::gsva(expr = assayDataElement(target_Data,
                                   elt = "log_q"),
                            gset.idx.list = immune_panels,
                            method = "ssgsea",
                            min.sz = 3,
                            max.sz = 500,
                            verbose = FALSE)

#subset target_data based on the genesets
geneSetObj <-
  NanoStringGeoMxSet(assayData = ssgsea_results,
                     phenoData = AnnotatedDataFrame(pData(target_Data)),
                     protocolData = protocolData(target_Data),
                     featureType = "GeneSet",
                     check = FALSE)

#make a new df with only the expression data of the genesetobject
pw_matrix<-assayDataElement(geneSetObj, elt = "exprs")

```


## Non-DE pathways

```{r, fig.height=15, fig.width=30}
#TLS
im_pathways_TLS <- pheatmap(ssgsea_results[,active_aois_TLS],
           scale = "row", 
           cluster_rows = F,
           show_rownames = TRUE, show_colnames = FALSE,
           annotation_names_row = F,
           gaps_row = c(12, 27),
           fontsize_row = 11,
           clustering_method = "complete",
           clustering_distance_cols = "correlation",
           clustering_distance_rows = "correlation",
           breaks = seq(-3, 3, 0.05),
           color = colorRampPalette(c("aquamarine", "black", "coral"))(120),
           annotation_colors = c(ann_c_TLS, ann_c_pt, ann_c_tissue, ann_c_maturity),
           cellheight = 12, cellwidth = 12, 
           border_color = "grey",
           main = "Pathway analysis TLS (non-DE)",
           annotation_col = pData(target_Data)[, c("maturity","tissue","patient", "region")])

save_pheatmap_pdf(x = im_pathways_TLS, file = paste0(figdir,"DSP1_TLS_non_DE_im_pathways_hm.pdf")) 

#TIL
im_pathways_TIL <- pheatmap(ssgsea_results[,active_aois_TIL],
           scale = "row", 
           cluster_rows = F,
           show_rownames = TRUE, show_colnames = FALSE,
           annotation_names_row = F,
           gaps_row = c(12, 27),
           fontsize_row = 11,
           clustering_method = "complete",
           clustering_distance_cols = "correlation",
           clustering_distance_rows = "correlation",
           breaks = seq(-3, 3, 0.05),
           color = colorRampPalette(c("aquamarine", "black", "coral"))(120),
           annotation_colors = c(ann_c_TIL, ann_c_pt, ann_c_tissue),
           cellheight = 12, cellwidth = 12, 
           border_color = "grey",
           main = "Pathway analysis TIL (non-DE)",
           annotation_col = pData(target_Data)[, c("tissue","patient", "region")])

save_pheatmap_pdf(x = im_pathways_TIL, file = paste0(figdir,"DSP1_TIL_non_DE_im_pathways_hm.pdf")) 
  
```

```{r fig.width=20, fig.height=15}
im_pathways_TLS <- pheatmap(ssgsea_results[,active_aois_TLS],
           scale = "row", 
           cluster_rows = T,
           cutree_cols = 4, cutree_rows = 3,
           show_rownames = TRUE, show_colnames = FALSE,
           annotation_names_row = F,
           fontsize_row = 11,
           clustering_method = "complete",
           #clustering_distance_cols = "correlation",
           #clustering_distance_rows = "correlation",
           breaks = seq(-3, 3, 0.05),
           color = colorRampPalette(c("aquamarine", "black", "coral"))(120),
           annotation_colors = c(ann_c_TLS, ann_c_pt, ann_c_tissue, ann_c_maturity),
           cellheight = 12, cellwidth = 12, 
           border_color = "grey",
           #main = "Pathway analysis TLS (non-DE)",
           annotation_col = pData(target_Data)[, c("maturity", "tissue","patient", "region")])

save_pheatmap_pdf(x = im_pathways_TLS, file = paste0(figdir,"DSP1_TLS_non_DE_im_pathways_with rowclust_hm.pdf")) 
```

# Chemokine signature
In all CD45 regions
```{r unsupervised chemokine, fig.height=10, fig.width=20}
genes <- rownames(exprs(target_Data))

# select genes from 12-gene cytokine signature
cyto.sig.all <- c("CCL19", "CCL21", "CXCL13", "CCL5", "CXCL9", "CXCL10", 
                  "CCL3", "CCL4", "CXCL11", "CCL2", "CCL18", "CCL8")
cyto.sig <- subset(genes, genes %in% cyto.sig.all)

# check which genes are present in dataset
setdiff(cyto.sig.all, cyto.sig)

#create a CD45 df with custom aoi names
data <- log2(assayDataElement(target_Data[cyto.sig,active_aois_CD45], elt = "q_norm"))
order <- sort(colnames(data))
data <- data[, order] #reorder columns alphabetically
colnames(data) <- ann$name[which(ann$segment == "CD45")]
ann_CD45 <- subset(ann, segment == "CD45")
rownames(ann_CD45) <- ann_CD45$name

#CD45
cytokine_signature_CD45 <- pheatmap(data,
         scale = "row", 
         angle_col = 90,
         main = "Cytokine signature of CD45 segment",
         show_rownames = T, show_colnames = F,
         clustering_method = "complete",
         clustering_distance_cols = "correlation",
         clustering_distance_rows = "correlation",
         cutree_cols = 3,cutree_rows = 2,
         breaks = seq(-3, 3, 0.05),
         color = colorRampPalette(c("aquamarine", "black", "coral"))(120),
         annotation_colors = c(ann_c_TLS_TIL, ann_c_pt, ann_c_tissue, ann_c_maturity),
         cellheight = 12, cellwidth = 12, 
         border_color = "grey",
         annotation_col = ann_CD45[, c("maturity", "tissue","patient", "region")])

save_pheatmap_pdf(x = cytokine_signature_CD45, file = paste0(figdir,"DSP1_cytokine_signature_CD45_hm.pdf")) 

#TLS
cytokine_signature_TLS <- pheatmap(log2(assayDataElement(target_Data[cyto.sig,active_aois_TLS], elt = "q_norm")),
         scale = "row", 
         angle_col = 90,
         main = "Cytokine signature of TLS",
         show_rownames = T, show_colnames = F,
         clustering_method = "average",
         clustering_distance_cols = "correlation",
         clustering_distance_rows = "correlation",
         cutree_cols = 2,cutree_rows = 1,
         breaks = seq(-3, 3, 0.05),
         color = colorRampPalette(c("aquamarine", "black", "coral"))(120),
         annotation_colors = c( ann_c_pt, ann_c_tissue, ann_c_maturity, ann_c_TLS, ann_c_maturity2),
         cellheight = 12, cellwidth = 12, 
         border_color = "grey",
         annotation_col = pData(target_Data)[, c("maturity2", "maturity", "tissue","patient", "region")])

save_pheatmap_pdf(x = cytokine_signature_TLS, file = paste0(figdir,"DSP1_cytokine_signature_TLS_hm.pdf"))  

#all regions
cytokine_signature_regions <- pheatmap(log2(assayDataElement(target_Data[cyto.sig,], elt = "q_norm")),
         scale = "row",
         angle_col = 90,
         main = "Cytokine signature of all regions",
         show_rownames = T, show_colnames = F,
         clustering_method = "complete",
         clustering_distance_cols = "correlation",
         clustering_distance_rows = "correlation",
         cutree_cols = 4,cutree_rows = 1,
         breaks = seq(-3, 3, 0.05),
         color = colorRampPalette(c("aquamarine", "black", "coral"))(120),
         annotation_colors = c(ann_c_pt, ann_c_tissue, ann_c_region, ann_c_maturity, ann_c_maturity2),
         cellheight = 12, cellwidth = 12,
         border_color = "grey",
         annotation_col = pData(target_Data)[, c("maturity2", "maturity", "tissue","patient", "region")])

save_pheatmap_pdf(x = cytokine_signature_regions, file = paste0(figdir,"DSP1_cytokine_signature_regions_hm.pdf")) 
```



# Correlation TLS maturity and TIL
prepare df for TLS
```{r mature_pt}
#assign maturity status per patient
ann_TLS$patient <- factor(ann_TLS$patient, levels=c('P1','P2','P3','P4','P5','P6','P7','P8','P9','P10'))
mature_pt <- unclass(table(ann_TLS$patient, ann_TLS$maturity))
times <- unclass(table(ann_TLS$patient))

a <- colnames(mature_pt)[apply(mature_pt,1,which.max)] #select most common state per pt
b <- rep(a, times = times) #repeat state times the number of samples per patient

ann_TLS$mature_pt <- b 
ann$mature_pt <- NA
ann$mature_pt[which(ann$region == "TLS")] <- b

#merge with props_TLS
props_TLS$mature_pt <- a
props_TLS$region <- "TLS" 

###
table(ann$patient, ann$maturity2)
mature <- c("P1","P2", "P5", "P6")
not_mature <- c("P3", "P7", "P9")

ann$mature_pt2 <- NA
ann$mature_pt2[which(ann$patient %in% mature)] <- "mature"
ann$mature_pt2[which(ann$patient %in% not_mature)] <- "non-mature"

ann_TLS <- subset(ann, region == "TLS")
ann_TLS$patient <- factor(ann_TLS$patient, levels=c('P1','P2','P3','P4','P5','P6','P7','P8','P9','P10'))

##

pData(target_Data)$mature_pt <- ann$mature_pt
pData(target_Data)$mature_pt2 <- ann$mature_pt2
pData(target_Data)$mature_pt2[which(pData(target_Data)$patient %in% c("P4", "P8", "P10"))] <- "ambiguous"
```



# Maturity selection figure
```{r warning = F}
mat_selection <- as.data.frame(unclass(table(ann_TLS$patient, ann_TLS$maturity))) %>%
  mutate(mat_pt = ann_TLS %>% 
          distinct(patient, mature_pt) %>% 
          select(mature_pt)) %>%
          select("LA", "early TLS", "PF-like TLS", "SF-like TLS", "mat_pt")

#in df
mat_selection$mat_pt2 <- NA
mat_selection$mat_pt2[which(rownames(mat_selection) %in% mature)] <- "mature"
mat_selection$mat_pt2[which(rownames(mat_selection) %in% not_mature)] <- "non-mature"
mat_selection$mat_pt2[which(rownames(mat_selection) %in% c("P4", "P8", "P10"))] <- "ambiguous"

levels <-  c( "ambiguous", "mature", "non-mature")
col_scale <- brewer.pal(max(mat_selection[,1:4]), "Greens")
ann_c_maturity2 <- list("mat_pt2" = c("non-mature" = "#CBB9F9", "mature" = "#9066FA", "ambiguous" = "white"))

row_ann <- HeatmapAnnotation(mat_pt = as.matrix(mat_selection[,6]), 
                             which = "row",
                             show_annotation_name = F,
                             col = list(mat_pt  = ann_c_maturity2[[1]],
                               annotation_legend_param = list(mat_pt = list(
                               title = "assigned state", 
                               title_position = "topleft",
                               labels = levels))),
                            width = unit(5, "mm"))

pdf(file = paste0(figdir, "DSP1_maturity_pt_assignment.pdf"), height = 5, width = 3.5)

Heatmap(mat_selection[1:4], 
        col = col_scale, 
        name = "no. of states",
        cluster_rows = F, 
        cluster_columns = F, 
        column_names_side = "top", 
        row_names_side = "right", 
        right_annotation = row_ann)

dev.off()

```

# IHC TILs boxplot / mature_pt

```{r}
TIL_table <- read_xlsx("cohort1/20231027_TLS_IHC_analyse.xlsx", sheet = "tumor_sum" )
TIL_table$patient <- factor(TIL_table$patient, levels = unique(ann$patient))
TIL_table$mature_pt <- factor(TIL_table$mature_pt, levels = c( "ambiguous", "non-mature", "mature"))

#calculations
TIL_table$CD3_p <- TIL_table$CD3 / TIL_table$total_cells * 100
TIL_table$CD8_p <- TIL_table$CD8 / TIL_table$total_cells * 100
TIL_table$FOXP3_p <- TIL_table$FOXP3 / TIL_table$total_cells * 100
TIL_table$CD163_p <- TIL_table$CD163 / TIL_table$total_cells * 100
TIL_table$KI67_p <- TIL_table$KI67 / TIL_table$total_cells * 100

df <- TIL_table %>%
  filter(mature_pt %in% c("non-mature", "mature")) %>%
  select(patient:mature_pt, CK:KI67) %>%
  pivot_longer(cols = CK:KI67, names_to = "markers", values_to = "counts")

stat_test <- df %>%
  group_by(markers) %>%
  wilcox_test(counts ~ mature_pt) %>%
  mutate(FDR = p.adjust(.$p, method = "fdr"))

for (marker in unique(df$markers)) {
print(
      ggplot(df[which(df$markers == marker),], aes(x = mature_pt, y = counts)) +
      geom_boxplot(aes(fill = mature_pt),
                     position = position_dodge(0.8), 
                     width = 0.6, 
                     fatten = 1) +
      geom_point() +
      scale_fill_manual(values = ann_c_maturity2[[1]][1:4]) +
      stat_pvalue_manual(stat_test %>% 
                           filter(marker %in% markers) %>% 
                           add_y_position, 
                         hide.ns = T, 
                         tip.length = 0) +
      theme_classic() +
      theme_bp()+
      theme(aspect.ratio = 1.5) +
      ylab("counts") +
      ggtitle(marker)
)
}
```
try the same with percentages instead of counts
```{r}
df <- TIL_table %>%
  filter(mature_pt %in% c("non-mature", "mature")) %>%
  select(patient:mature_pt, CD3_p:KI67_p) %>%
  pivot_longer(cols = CD3_p:KI67_p, names_to = "markers", values_to = "counts")

stat_test <- df %>%
  group_by(markers) %>%
  wilcox_test(counts ~ mature_pt) %>%
  mutate(FDR = p.adjust(.$p, method = "fdr"))

for (marker in unique(df$markers)) {
print(
      ggplot(df[which(df$markers == marker),], aes(x = mature_pt, y = counts)) +
      geom_boxplot(aes(fill = mature_pt),
                     position = position_dodge(0.8), 
                     width = 0.6, 
                     fatten = 1) +
      geom_point() +
      scale_fill_manual(values = ann_c_maturity2[[1]][1:4]) +
      stat_pvalue_manual(stat_test %>% 
                           filter(marker %in% markers) %>% 
                           add_y_position, 
                         hide.ns = T, 
                         tip.length = 0) +
      theme_classic() +
      theme_bp()+
      theme(aspect.ratio = 1.5) +
      ylab("% postive cells") +
      ggtitle(marker)
)
}
```


## LMM-DE analysis TLS: mature vs non-mature
```{r}
target_Data_TLS <- subset(geneSetObj,select = pData(geneSetObj)[["region"]] == "TLS")
target_Data_TLS <- subset(target_Data_TLS, select = phenoData(target_Data_TLS)[["maturity2"]] %in% c("non-mature", "mature"))

pData(target_Data_TLS)$dfRegion <- as.factor(pData(target_Data_TLS)$region)
pData(target_Data_TLS)$dfPatient <- factor(pData(target_Data_TLS)$patient,
                                       levels=c("P1","P2","P3","P5","P6","P7","P9"))
pData(target_Data_TLS)$dfMature <- factor(pData(target_Data_TLS)$maturity2, levels = c("mature", "non-mature"))

  mixedOutmc <-
    mixedModelDE(target_Data_TLS,
                 elt = "exprs",
                 modelFormula = ~ dfMature + (1 | dfPatient),
                 groupVar = "dfMature",
                 multiCore = FALSE,
                 pAdjust = NULL
                 )
  
  lsmeans <- lapply(mixedOutmc, function(x) {x$lsmeans})
  anova <- unlist(lapply(mixedOutmc, function(x) {x$anova}))
  b <- lapply(lsmeans, function(x) {x[1]})
  c <- lapply(lsmeans, function(x) {x[2]})
  lmm_ssgsea_results <- cbind.data.frame(unlist(b), unlist(c))
  names(lmm_ssgsea_results) <- c('LogFC', 'Pr(>|t|)')
 
  ssgsea_dfs <- rownames(lmm_ssgsea_results)
  lmm_ssgsea_results$Pathway <- rownames(lmm_ssgsea_results)
  lmm_ssgsea_results$FDR <- p.adjust(lmm_ssgsea_results$`Pr(>|t|)`, method = "fdr")
  lmm_ssgsea_results <- lmm_ssgsea_results[, c("Pathway", "LogFC",
                                     "Pr(>|t|)", "FDR")]
```

```{r fig.height=10, fig.width=20}
ann$aois <- rownames(ann)
ann_c_maturity2 <- list("maturity2" = c("non-mature" = "#CBB9F9", "mature" = "#9066FA"))

pw_matrix<-assayDataElement(target_Data_TLS, elt = "exprs")
order <- c(ann$aois[which(ann$maturity2 == "mature" & ann$region == "TLS")], 
           ann$aois[which(ann$maturity2 == "non-mature"& ann$region == "TLS")])
pw_matrix <- pw_matrix[,order]

active_pw<-filter(lmm_ssgsea_results, `Pr(>|t|)` < 0.05 )$Pathway

DE_im_pathways <- pheatmap(pw_matrix[active_pw,],
           scale = "row", 
           show_rownames = TRUE, show_colnames = F, 
           gaps_col = 30,
           annotation_names_row = F,
           fontsize_row = 11,
           cluster_cols = F,
           # clustering_method = "complete",
           # clustering_distance_cols = "euclidean",
           clustering_distance_rows = "euclidean",
           cutree_cols = 2, cutree_rows = 2,
           breaks = seq(-3, 3, 0.05),
           color = colorRampPalette(c("aquamarine", "black", "coral"))(120),
           cellheight = 12, cellwidth = 12, 
           main = "mature vs non-mature TLS, p < 0.05",
           border_color = "grey",
           annotation_colors = c(ann_c_TLS_TIL, ann_c_pt, ann_c_tissue, ann_c_panels, ann_c_maturity, ann_c_maturity2),
           annotation_col = pData(target_Data_TLS)[, c("maturity2", "maturity","tissue","patient", "region")])

save_pheatmap_pdf(x = DE_im_pathways, file = paste0(figdir,"DSP1_CD45_DE_im_pathways_p0.05.pdf"))

active_pw<-filter(lmm_ssgsea_results, FDR < 0.05 )$Pathway

DE_im_pathways <- pheatmap(pw_matrix[active_pw,],
           scale = "row", 
           show_rownames = TRUE, show_colnames = F, 
           gaps_col = 30,
           annotation_names_row = F,
           fontsize_row = 11,
           cluster_cols = F,
           # clustering_method = "complete",
           # clustering_distance_cols = "euclidean",
           clustering_distance_rows = "euclidean",
           cutree_cols = 2, cutree_rows = 2,
           breaks = seq(-3, 3, 0.05),
           color = colorRampPalette(c("aquamarine", "black", "coral"))(120),
           cellheight = 12, cellwidth = 12, 
           main = "mature vs non-mature TLS, FDR < 0.05",
           border_color = "grey",
           annotation_colors = c(ann_c_TLS_TIL, ann_c_pt, ann_c_tissue, ann_c_panels, ann_c_maturity, ann_c_maturity2),
           annotation_col = pData(target_Data_TLS)[, c("maturity2", "maturity","tissue","patient", "region")])

save_pheatmap_pdf(x = DE_im_pathways, file = paste0(figdir,"DSP1_CD45_DE_im_pathways_FDR0.05.pdf"))
```

# Dotplots TIL vs mature patients

## DE pathways
This uses a selected gene set object (ssgsea_results from gsva analysis)
```{r pathways}
aois <- rownames(ann)[which(ann$region == "TIL" & ann$patient %in% c("P1", "P5", "P6", "P2","P3", "P7", "P9"))]

ssgsea_results <- GSVA::gsva(expr = assayDataElement(target_Data,
                                   elt = "log_q"),
                            gset.idx.list = immune_panels,
                            method = "ssgsea",
                            min.sz = 3,
                            max.sz = 500,
                            verbose = FALSE)

#subset target_data based on the genesets
geneSetObj <-
  NanoStringGeoMxSet(assayData = ssgsea_results,
                     phenoData = AnnotatedDataFrame(pData(target_Data)),
                     protocolData = protocolData(target_Data),
                     featureType = "GeneSet",
                     check = FALSE)

geneSetObj_TIL <-subset(geneSetObj, select = phenoData(geneSetObj)[["region"]] == "TIL")
geneSetObj_TIL <-subset(geneSetObj_TIL, select = phenoData(geneSetObj_TIL)[["mature_pt2"]] %in% c("non-mature", "mature"))

pData(geneSetObj_TIL)$dfRegion <- as.factor(pData(geneSetObj_TIL)$region)
pData(geneSetObj_TIL)$dfPatient <- factor(pData(geneSetObj_TIL)$patient,
                                       levels=c("P1","P2","P3","P5","P6","P7","P9"))
pData(geneSetObj_TIL)$dfMature <- factor(pData(geneSetObj_TIL)$mature_pt2, levels = c("non-mature", "mature"))
pData(geneSetObj_TIL)[["slide"]] <- factor(pData(geneSetObj_TIL)[["slide name"]])

   mixedOutmc <-
    mixedModelDE(geneSetObj_TIL,
                 elt = "exprs",
                 modelFormula = ~ dfMature + (1 | dfPatient),
                 groupVar = "dfMature",
                 multiCore = FALSE,
                 pAdjust = NULL
                 )
  
  lsmeans <- lapply(mixedOutmc, function(x) {x$lsmeans})
  anova <- unlist(lapply(mixedOutmc, function(x) {x$anova}))
  b <- lapply(lsmeans, function(x) {x[1]})
  c <- lapply(lsmeans, function(x) {x[2]})
  lmm_ssgsea_results <- cbind.data.frame(unlist(b), unlist(c))
  names(lmm_ssgsea_results) <- c('LogFC', 'Pr(>|t|)')
 
  ssgsea_dfs <- rownames(lmm_ssgsea_results)
  lmm_ssgsea_results$Pathway <- rownames(lmm_ssgsea_results)
  lmm_ssgsea_results$FDR <- p.adjust(lmm_ssgsea_results$`Pr(>|t|)`, method = "fdr")
  lmm_ssgsea_results <- lmm_ssgsea_results[, c("Pathway", "LogFC",
                                     "Pr(>|t|)", "FDR")]

lmm_ssgsea_results %>% filter(`Pr(>|t|)` < 0.05)

pw_matrix<-assayDataElement(geneSetObj_TIL, elt = "exprs")
```

```{r}
goi <- lmm_ssgsea_results %>% filter(`Pr(>|t|)` < 0.05) %>% rownames

lmm_ssgsea_results %>% 
  as.data.frame %>% 
    remove_rownames %>% 
      mutate(`-log10 pval` = -log10(`Pr(>|t|)`)) %>%
        filter(Pathway %in% goi) %>%
          arrange(`-log10 pval`) %>%
            mutate(dir = factor(ifelse(LogFC > 0, "down", "up"), 
                                levels = c("up",  "down"))) %>%  #up in mature, down in not-mature
              mutate(Pathway = factor(Pathway, levels = rev(.$Pathway))) %>%
                ggplot(aes(x = Pathway, y = `Pr(>|t|)`)) +
                geom_segment(aes(x = Pathway, 
                             xend = Pathway, 
                             y = 0, 
                             yend = `Pr(>|t|)`,
                             color = dir),   
                             linewidth = 1) + 
                geom_hline(yintercept = c(0.05, 0.01), lty = "dashed", color = "black") +
                geom_point(aes(color = dir, size = LogFC)) +
                theme_classic() +
                theme(aspect.ratio = 0.5,
                      axis.title.y = element_blank(),
                      axis.title.x = element_text(face = "bold"),
                       plot.title = element_text(hjust = 0.5, face = "bold")) +
                coord_flip() +
                ylab("p value") +
                labs(size = "log2 FC", color = "direction") +
                ggtitle("DE pathways, p > 0.05")

ggsave("DSP1_Spearman_matVSnotmat_TIL_pathways.pdf", device = "pdf", height = 4, width = 4, path = figdir)
```

```{r eval = F}

active_pw<-filter(lmm_ssgsea_results, `Pr(>|t|)` < 0.05 )$Pathway

order <- c(ann$aois[which(ann$mature_pt2 == "mature" & ann$region == "TIL")], 
           ann$aois[which(ann$mature_pt2 == "non-mature"& ann$region == "TIL")])

pheatmap(pw_matrix[active_pw,order],
           scale = "row", 
           show_rownames = TRUE, show_colnames = F,
           annotation_names_row = F,
           fontsize_row = 11,
           clustering_method = "complete",
           cluster_cols = F,
           #clustering_distance_cols = "correlation",
           clustering_distance_rows = "correlation",
           cutree_cols = 2, cutree_rows = 2,
           breaks = seq(-3, 3, 0.05),
           color = colorRampPalette(c("aquamarine", "black", "coral"))(120),
           cellheight = 12, cellwidth = 12, 
           border_color = "grey",
           #main ="LMM DE: TLS vs TIL pathway analysis (FDR < 0.05)",
          annotation_colors = c(ann_c_TIL, ann_c_pt, ann_c_tissue, ann_c_panels, ann_c_maturity, ann_c_mature_pt2),
           annotation_col = pData(target_Data)[which( pData(target_Data)$region == "TIL"), c("mature_pt2","tissue", "region")])

#save_pheatmap_pdf(x = DE_im_pathways, file = paste0(figdir,"DSP1_CD45_DE_im_pathways_with maturityANN_hm_supervised.pdf")) 
```


## DE genes
This uses an object with all genes
```{r}
target_Data_TIL <- subset(target_Data,select = pData(target_Data)[["region"]] == "TIL")
target_Data_TIL <-subset(target_Data_TIL, select = phenoData(target_Data_TIL)[["mature_pt2"]] %in% c("non-mature", "mature"))

pData(target_Data_TIL)$dfRegion <- as.factor(pData(target_Data_TIL)$region)
pData(target_Data_TIL)$dfPatient <- factor(pData(target_Data_TIL)$patient,
                                       levels=c("P1","P2","P3","P5","P6","P7","P9"))
pData(target_Data_TIL)$dfMature <- factor(pData(target_Data_TIL)$mature_pt2, levels = c("mature", "non-mature"))

    mixedOutmc <-
        mixedModelDE(target_Data_TIL,
                     elt = "log_q",
                     modelFormula = ~ dfMature + (1 | dfPatient),
                     groupVar = "dfMature",
                     nCores = parallel::detectCores(),
                     multiCore = TRUE)
    
    # format results as data.frame
    lmm_results <- do.call(rbind, mixedOutmc["lsmeans", ])
    tests <- rownames(lmm_results)
    lmm_results <- as.data.frame(lmm_results)
    lmm_results$Contrast <- tests
    
    # use lapply in case you have multiple levels of your test factor to
    # correctly associate gene name with it's row in the results table
    lmm_results$Gene <- 
        unlist(lapply(colnames(mixedOutmc),
                      rep, nrow(mixedOutmc["lsmeans", ][[1]])))
    #lmm_results$Subset <- status
    lmm_results$FDR <- p.adjust(lmm_results$`Pr(>|t|)`, method = "fdr")
    lmm_results <- lmm_results[, c("Gene", "Contrast", "Estimate", 
                         "Pr(>|t|)", "FDR")]
    
#lmm_results %>% filter(`Pr(>|t|)` < 0.05) %>% remove_rownames()
#negative estimate indicates higher expression in second group in contrast
#thus postive is up in non-mature, negative up in mature

```


```{r}
goi <- immune_panels[["Cytotoxicity"]]

line <- 0.05

lmm_results %>% 
  as.data.frame %>% 
    remove_rownames %>% 
      mutate(`-log10 pval` = -log10(`Pr(>|t|)`)) %>%
        filter(Gene %in% goi) %>%
          arrange(`-log10 pval`) %>%
            mutate(dir = factor(ifelse(Estimate > 0, "down", "up"), 
                                levels = c("up",  "down"))) %>%  #up in mature, down in not-mature
              mutate(Gene = factor(Gene, levels = rev(.$Gene))) %>%
                ggplot(aes(x = Gene, y = `Pr(>|t|)`)) +
                geom_segment(aes(x = Gene, 
                             xend = Gene, 
                             y = 0, 
                             yend = `Pr(>|t|)`,
                             color = dir),   
                             linewidth = 1) + 
                geom_hline(yintercept = line, lty = "dashed", color = "black") +
                geom_point(aes(color = dir, size = Estimate)) +
                theme_classic() +
                theme(aspect.ratio = 1.5,
                      axis.title.y = element_blank(),
                      axis.title.x = element_text(face = "bold"),
                       plot.title = element_text(hjust = 0.5, face = "bold")) +
                coord_flip() +
                ylab("p value") +
                labs(size = "log2 FC", color = "direction") +
                ggtitle("cytotoxicity")

ggsave("DSP1_Spearman_matVSnotmat_TIL_genes_cytox.pdf", device = "pdf", height = 4, width = 4, path = figdir)
```

```{r eval = F}
data <- assayDataElement(target_Data_TIL[goi, ], elt = "log_q")

order <- c(ann$aois[which(ann$mature_pt2 == "mature" & ann$region == "TIL")], 
           ann$aois[which(ann$mature_pt2 == "non-mature"& ann$region == "TIL")])

pheatmap(data[,order],
           scale = "row", 
           show_rownames = TRUE, show_colnames = F,
           annotation_names_row = F,
           fontsize_row = 11,
           clustering_method = "complete",
           cluster_cols = F,
           #clustering_distance_cols = "correlation",
           clustering_distance_rows = "correlation",
           cutree_cols = 2, cutree_rows = 2,
           breaks = seq(-3, 3, 0.05),
           color = colorRampPalette(c("aquamarine", "black", "coral"))(120),
           cellheight = 12, cellwidth = 12, 
           border_color = "grey",
           #main ="LMM DE: TLS vs TIL pathway analysis (FDR < 0.05)",
          annotation_colors = c(ann_c_TIL, ann_c_pt, ann_c_tissue, ann_c_panels, ann_c_maturity, ann_c_mature_pt2),
           annotation_col = pData(target_Data)[which( pData(target_Data)$region == "TIL"), c("mature_pt2","tissue", "region")])

#save_pheatmap_pdf(x = DE_im_pathways, file = paste0(figdir,"DSP1_CD45_DE_im_pathways_with maturityANN_hm_supervised.pdf")) 
```

```{r}
goi <- immune_panels[["Other Interleukin Signaling"]]
line <- 0.05

lmm_results %>% 
  as.data.frame %>% 
    remove_rownames %>% 
      mutate(`-log10 pval` = -log10(`Pr(>|t|)`)) %>%
        filter(Gene %in% goi) %>% 
        filter(`Pr(>|t|)` < 0.05) %>%
          arrange(`-log10 pval`) %>%
            mutate(dir = factor(ifelse(Estimate > 0, "down", "up"), 
                                levels = c("up",  "down"))) %>%  #up in mature, down in not-mature
              mutate(Gene = factor(Gene, levels = rev(.$Gene))) %>%
                ggplot(aes(x = Gene, y = `Pr(>|t|)`)) +
                geom_segment(aes(x = Gene, 
                             xend = Gene, 
                             y = 0, 
                             yend = `Pr(>|t|)`,
                             color = dir),   
                             linewidth = 1) + 
                #geom_hline(yintercept = line, lty = "dashed", color = "black") +
                geom_point(aes(color = dir, size = Estimate)) +
                theme_classic() +
                theme(aspect.ratio = 1.5,
                      axis.title.y = element_blank(),
                      axis.title.x = element_text(face = "bold"),
                       plot.title = element_text(hjust = 0.5, face = "bold")) +
                coord_flip() +
                ylab("p value") +
                labs(size = "log2 FC", color = "direction") +
                ggtitle("other interleukin signaling")

ggsave("DSP1_Spearman_matVSnotmat_TIL_genes_ILsig.pdf", device = "pdf", height = 4, width = 4, path = figdir)
```
```{r}
goi <- lmm_results %>% filter(`Pr(>|t|)` < 0.01) %>% select(Gene) %>% unlist %>% as.vector

line <- 0.05

lmm_results %>% 
  as.data.frame %>% 
    remove_rownames %>% 
      mutate(`-log10 pval` = -log10(`Pr(>|t|)`)) %>%
        filter(Gene %in% goi) %>% 
          arrange(`-log10 pval`) %>%
            mutate(dir = factor(ifelse(Estimate < 0, "down", "up"), 
                                levels = c("up",  "down"))) %>%  #up in mature, down in not-mature
              mutate(Gene = factor(Gene, levels = rev(.$Gene))) %>%
                ggplot(aes(x = Gene, y = `Pr(>|t|)`)) +
                geom_segment(aes(x = Gene, 
                             xend = Gene, 
                             y = 0, 
                             yend = `Pr(>|t|)`,
                             color = dir),   
                             linewidth = 1) + 
                geom_hline(yintercept = c(0.01, 0.005, 0.001), lty = "dashed", color = "black") +
                geom_point(aes(color = dir, size = Estimate)) +
                theme_classic() +
                theme(axis.title.y = element_blank(),
                      axis.title.x = element_text(face = "bold"),
                       plot.title = element_text(hjust = 0.5, face = "bold")) +
                coord_flip() +
                labs(size = "log2 FC", color = "direction") +
                ylab("p value") +
                ggtitle("DE genes, p < 0.01")

ggsave("DSP1_Spearman_matVSnotmat_TIL_genes_p0.05.pdf", device = "pdf", height = 7, width = 4, path = figdir)
```

# Correlation TLS-subsets with TIL-subsets
prepare df for TIL and merge with TLS
```{r}
#merge with props_TIL
props_TIL$mature_pt2 <- mat_selection$mat_pt2
props_TIL$region <- "TIL" 
props_TLS$mature_pt2 <- mat_selection$mat_pt2

```

## mature pt
```{r}
means <- colMeans(props_TLS_all[,-1])
index <- names(means[which(means > 0.02)])
#index <- means[-7]

TLS_mat <- props_TLS[which(props_TLS$mature_pt2 == "mature"), index]
TIL_mat <- props_TIL[which(props_TIL$mature_pt2 == "mature"), index]

#substitude zeros with very low values
TIL_mat[TIL_mat == 0] <- 0.00000001
TIL_mat[2,9] <-  0.00000002
TIL_mat[2,7] <-  0.00000002

cor_mat <- cor(x = TLS_mat, y = TIL_mat)

cor_test_mat <- corr.test(x = TLS_mat, y = TIL_mat, method = "spearman")$p

pdf(file = paste0(figdir, "DSP1_corrplot subsets matTLS and TIL.pdf"))

corrplot(corr = cor_mat, 
         method = "ellipse",
         tl.col = "black",
         tl.cex = 1.5, 
         pch.col = "white",
         p.mat = cor_test_mat,
         sig.level = 0.05,
         insig = "label_sig",
         mar = c(2, 2, 2, 2))

dev.off()

```
## immature pt

```{r}
means <- colMeans(props_TLS_all[,-1])
index <- names(means[which(means > 0.02)])

TLS_mat <- props_TLS[which(props_TLS$mature_pt2 == "non-mature"), index]
TIL_mat <- props_TIL[which(props_TIL$mature_pt2 == "non-mature"), index]

#substitude zeros with very low values
TIL_mat[TIL_mat == 0] <- 0.0000001

cor_mat <- cor(x = TLS_mat, y = TIL_mat)

cor_test_mat <- corr.test(x = TLS_mat, y = TIL_mat, method = "spearman")$p

pdf(file = paste0(figdir, "DSP1_corrplot subsets imTLS and TIL.pdf"))

corrplot(corr = cor_mat, 
         method = "ellipse",
         tl.col = "black",
         tl.cex = 1.5, 
         pch.col = "white",
         p.mat = cor_test_mat,
         sig.level = 0.05,
         insig = "label_sig",
         mar = c(2, 2, 2, 2))

dev.off()

```

## all pt
```{r fig.width=10, fig.height=10}
means <- colMeans(props_TLS_all[,-1])
index <- names(means[which(means > 0.02)])

TLS_mat <- props_TLS[, index]
TIL_mat <- props_TIL[, index]

cor_mat <- cor(x = TLS_mat, y = TIL_mat)

cor_test_mat <- corr.test(x = TLS_mat, y = TIL_mat, method = "spearman")$p

pdf(file = paste0(figdir, "DSP1_corrplot subsets TLS and TIL_all pt.pdf"))

corrplot(corr = cor_mat, 
         method = "ellipse",
         tl.col = "black",
         tl.cex = 1.5, 
         pch.col = "white",
         p.mat = cor_test_mat,
         sig.level = 0.05,
         insig = "label_sig",
         mar = c(2, 2, 2, 2))

dev.off()
 
#cor_mat[cor_mat > -0.6 & cor_mat < 0.6] <- NA
#cor_mat


```

# correlation chemokines in 12-ck en immune subsets in TLS en TIL
spearman correlation was performed but most data seemed too significant for the low correlations. 
```{r fig.height=10, fig.width=10}
cytokine_mat <- log2(assayDataElement(target_Data[cyto.sig,active_aois_TLS], elt = "q_norm"))
order <- sort(colnames(cytokine_mat))
cytokine_mat <- cytokine_mat[, order]
colnames(cytokine_mat) <- ann$name[which(ann$region == "TLS")]
cytokine_mat <- t(cytokine_mat) %>% as.data.frame 

prop_mat <- props_TLS_all
rownames(prop_mat) <- prop_mat[,1]
prop_mat <- prop_mat[,-1]
prop_mat %<>% select(all_of(coi))
```

```{r fig.height=5, fig.width=5, eval = F}
#to produce scatterplots of data
for (cells in colnames(prop_mat)) {
  for (chemo in colnames(cytokine_mat)) {
    
    a <- cbind(prop = prop_mat[,as.character(cells)], 
               chemo = cytokine_mat[,as.character(chemo)]) %>% 
      as.data.frame
    
    cor_coef <- round(cor_mat[chemo, cells], digits = 2)
    cor_pval <- round(cor_test_mat[chemo, cells], digits = 4)
    
    if(cor_coef > -0.6 & cor_coef < 0.6 | cor_pval > 0.05 ) next
    
      print(
        ggplot(a, aes(x = chemo, y = prop)) +
        geom_point() +
        geom_smooth(color = "#9e42f5", method = "lm", se = F) +
        ggtitle(paste(cells, chemo, "rho =", cor_coef, "p =", cor_pval)) + 
        theme_bw() +
        theme(plot.title = element_text(hjust = 0.5, face = "bold"), 
              legend.title = element_blank(), legend.position = "none",
              aspect.ratio = 1,
              axis.title = element_text(face = "bold")) +
        xlab("log2 expression") +
        ylab("proportion")
      )
        
    ggsave(paste0("DSP1_Spearman_TLS-chemo_TIL-subs_",cells,"_", chemo, ".pdf"), 
                  device = "pdf", path = paste0(figdir, "scatterplots"))
  }
}

#are the correlations explained by the spatial decon matrix?
# current_cell_profile %>% 
#   as.data.frame %>%
#   mutate(across(where(is.numeric), round, 3)) %>%
#   rownames_to_column("genes") %>%
#   filter(genes %in% c("CCL5", "CCL4", "CCL2", "CXCL9"))

#it explains the correlation between CCL5 and CD8 Tmemory, but not the other correlations

```

## scatterplots
```{r}
cytokine_mat <- log2(assayDataElement(target_Data[cyto.sig,active_aois_TLS], elt = "q_norm"))
order <- sort(colnames(cytokine_mat))
cytokine_mat <- cytokine_mat[, order]
colnames(cytokine_mat) <- ann$name[which(ann$region == "TLS")]
cytokine_mat <- t(cytokine_mat) %>% as.data.frame 

prop_mat <- props_TLS_all
rownames(prop_mat) <- prop_mat[,1]
prop_mat <- prop_mat[,-1]
prop_mat %<>% select(all_of(coi))

cor_mat <- cor(x = cytokine_mat, y = prop_mat)
cor_test_mat <- corr.test(x = cytokine_mat, y = prop_mat, method = "spearman")$p

a <- cbind(prop = prop_mat[,"macrophages"], 
           chemo = cytokine_mat[,"CCL2"]) %>% 
  as.data.frame

cor_coef <- round(cor_mat["CCL2", "macrophages"], digits = 2)
cor_pval <- round(cor_test_mat["CCL2", "macrophages"], digits = 10)

ggplot(a, aes(x = chemo, y = prop)) +
  geom_point() +
  geom_smooth(color = "#9e42f5", method = "lm", se = F) +
  annotate("text", x = 5, y = 0.25, hjust = 0, label = paste0("rho =", cor_coef)) +
  annotate("text", x= 5, y = 0.23, hjust = 0, label = substitute(paste(italic("p"), " < 0.001"))) +
  ggtitle(paste("macrophages - CCL2")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), 
        legend.title = element_blank(), legend.position = "none",
        aspect.ratio = 1,
        axis.title = element_text(face = "bold")) +
  xlab("log2 expression") +
  ylab("proportion")

ggsave("DSP1_Spearman_macro_CCL2.pdf", device = "pdf", path = figdir)

################
a <- cbind(prop = prop_mat[,"CD8 T memory"], 
           chemo = cytokine_mat[,"CCL4"]) %>% 
  as.data.frame

cor_coef <- round(cor_mat["CCL4", "CD8 T memory"], digits = 2)
cor_pval <- round(cor_test_mat["CCL4", "CD8 T memory"], digits = 10)

ggplot(a, aes(x = chemo, y = prop)) +
  geom_point() +
  geom_smooth(color = "#9e42f5", method = "lm", se = F) +
  annotate("text", x = 5, y = 0.38, hjust = 0, label = paste0("rho =", cor_coef)) +
  annotate("text", x= 5, y = 0.36, hjust = 0, label = substitute(paste(italic("p"), " < 0.001"))) +
  ggtitle(paste("CD8 T memory - CCL4")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), 
        legend.title = element_blank(), legend.position = "none",
        aspect.ratio = 1,
        axis.title = element_text(face = "bold")) +
  xlab("log2 expression") +
  ylab("proportion")

ggsave("DSP1_Spearman_Cd8Tmem_CCL4.pdf", device = "pdf", path = figdir)

################

a <- cbind(prop = prop_mat[,"CD8 T memory"], 
           chemo = cytokine_mat[,"CXCL9"]) %>% 
  as.data.frame

cor_coef <- round(cor_mat["CXCL9", "CD8 T memory"], digits = 2)
cor_pval <- round(cor_test_mat["CXCL9", "CD8 T memory"], digits = 10)

ggplot(a, aes(x = chemo, y = prop)) +
  geom_point() +
  geom_smooth(color = "#9e42f5", method = "lm", se = F) +
  annotate("text", x = 3.5, y = 0.38, hjust = 0, label = paste0("rho =", cor_coef)) +
  annotate("text", x= 3.5, y = 0.36, hjust = 0, label = substitute(paste(italic("p"), " < 0.001"))) +
  ggtitle(paste("CD8 T memory - CXCL9")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), 
        legend.title = element_blank(), legend.position = "none",
        aspect.ratio = 1,
        axis.title = element_text(face = "bold")) +
  xlab("log2 expression") +
  ylab("proportion")

ggsave("DSP1_Spearman_Cd8Tmem_CXCL9.pdf", device = "pdf", path = figdir)

```

## mmlm TLS subsets & TLS chemokines
we try to do a similar analysis using a multivariate multiple linear model
TLS chemokine correlation with TLS subsets
```{r warning = F, include = F}
prop_mat %<>% as.matrix
cytokine_mat %<>% as.matrix

test <- lm(prop_mat ~ cytokine_mat)
summary(test)

#estimates mmlm
heatmap_estimate<- test %>% 
  tidy %>% 
  filter(term != "(Intercept)") %>% 
  select(response, term, estimate) %>% 
  pivot_wider(names_from = term, values_from = estimate) %>% 
  column_to_rownames(var = "response") %>% 
  scale()
colnames(heatmap_estimate) <- cyto.sig

#pvalues mmlm
heatmap_p <- test %>% 
  tidy %>% 
  filter(term != "(Intercept)") %>% 
  select(response, term, `p.value`) %>% 
  pivot_wider(names_from = term, values_from = `p.value`) %>% 
  column_to_rownames(var = "response") 
colnames(heatmap_p) <- cyto.sig

#manual annotation of pvalues in heatmap
panno_function = function(j, i, x, y, width, height, fill) {if(heatmap_p[i, j] <= 0.05) grid.text(round(heatmap_p[i, j], digits = 3), x, y, gp = gpar(fontsize = 10, font="bold"))} 

pdf(file = paste0(figdir, "DSP1_mmlm TLS chemokines and subsets_hm.pdf"), height = 7, width = 8)

Heatmap(heatmap_estimate, cell_fun = panno_function)

dev.off()
```

# Correlation TLS chemokines and TIL subsets
TLS chemokine correlation with TIL subsets
first calculate mean per patient

```{r, fig.height=20, fig.width=20}
patients <- unique(ann$patient)

cytokine_mat <- data.frame(row.names = cyto.sig)

for (pt in patients) {
  
  index <- rownames(ann)[which(ann$region == "TLS" & ann$patient == as.character(pt))]
  n <- length(index)
 
   cytokines <- log2(assayDataElement(target_Data[cyto.sig, index], elt = "q_norm")) %>%
    rowSums/n
   
  cytokines <- as.data.frame(cytokines)
  colnames(cytokines) <- pt
  
  cytokine_mat <- cbind(cytokine_mat, cytokines)
}

prop_mat <- data.frame(row.names = prettynames)

for (pt in patients) {
  
  index <- rownames(ann)[which(ann$region == "TIL" & ann$patient == as.character(pt))]

  n <- length(index)
  
  props <- props_TIL_all %>% 
    filter(name %in% ann$name[which(ann$region == "TIL" & ann$patient == as.character(pt))]) %>%
    select(macrophages:Treg) %>%
    colSums/n
  
  props <- as.data.frame(props)
  colnames(props) <- pt
  
  prop_mat <- cbind(prop_mat, props)
}

prop_mat %<>% t
cytokine_mat %<>% t

```

```{r fig.height=5, fig.width=5}
#to remove low abundance subsets
means <- colMeans(prop_mat)
means <- names(means[which(means > 0.03)])

#to produce scatterplots of data
for (cells in colnames(prop_mat[,means])) {
  for (chemo in colnames(cytokine_mat)) {
    
    a <- cbind(prop = prop_mat[,as.character(cells)], 
               chemo = cytokine_mat[,as.character(chemo)]) %>% 
      as.data.frame
    
    cor_coef <- round(cor_mat[chemo, cells], digits = 2)
    cor_pval <- round(cor_test_mat[chemo, cells], digits = 4)
    
    if(cor_coef > -0.6 & cor_coef < 0.6 | cor_pval > 0.05 ) next
    
      print(
        ggplot(a, aes(x = chemo, y = prop)) +
        geom_point() +
        geom_smooth(color = "#9e42f5", method = "lm", se = F) +
        ggtitle(paste(cells, chemo, "rho =", cor_coef, "p =", cor_pval)) + 
        theme_bw() +
        theme(plot.title = element_text(hjust = 0.5, face = "bold"), 
              legend.title = element_blank(), legend.position = "none",
              aspect.ratio = 1,
              axis.title = element_text(face = "bold")) +
        xlab("log2 expression") +
        ylab("proportion")
      )
        
    #ggsave(paste0("DSP1_Spearman_TLS-chemo_TIL-subs_",cells,"_", chemo, ".pdf"), 
    #              device = "pdf", path = paste0(figdir, "scatterplots"))
  }
}

#check if correlations are due to spatial decon matrix
current_cell_profile %>%
  as.data.frame %>%
  mutate(across(where(is.numeric), round, 3)) %>%
  rownames_to_column("genes") %>%
  filter(genes %in% cyto.sig)

# potential fake associations:
# memory CD4/CD8 T cells / CCL5
# MQ CCL18, and CXCL9 and CCL5 (low in matrix)
# CD8 T memory CCL4 (very low in matrix)

# potential real associations (0 in matrix)
# B naive CCL18
# CD4 T naive CXCL10 (-)
# CD4 T mem CCL18 (-), CXCL9 (+)
# CD8 T memory CXCL10, CXCL13


```
## scatterplots
```{r}
a <- cbind(prop = prop_mat[,"B naive"], 
           chemo = cytokine_mat[,"CCL18"]) %>% 
  as.data.frame

cor_coef <- round(cor_mat["CCL18", "B naive"], digits = 2)
cor_pval <- round(cor_test_mat["CCL18", "B naive"], digits = 3)

ggplot(a, aes(x = chemo, y = prop)) +
  geom_point() +
  geom_smooth(color = "#9e42f5", method = "lm", se = F) +
  annotate("text", x = 5.2, y = 0.115, hjust = 0, label = paste0("rho = ", cor_coef)) +
  annotate("text", x= 5.2, y = 0.105, hjust = 0, label = substitute(paste(italic("p"), " = 0.038"))) +
  ggtitle(paste("B naive - CCL18")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), 
        legend.title = element_blank(), legend.position = "none",
        aspect.ratio = 1,
        axis.title = element_text(face = "bold")) +
  xlab("log2 expression") +
  ylab("proportion")

ggsave("DSP1_Spearman_TIL-Bnaive_TLS-CCL18.pdf", device = "pdf", height = 4, width = 4, path = figdir)

########

a <- cbind(prop = prop_mat[,"CD4 T memory"], 
           chemo = cytokine_mat[,"CCL18"]) %>% 
  as.data.frame

cor_coef <- round(cor_mat["CCL18", "CD4 T memory"], digits = 2)
cor_pval <- round(cor_test_mat["CCL18", "CD4 T memory"], digits = 3) %>% print

ggplot(a, aes(x = chemo, y = prop)) +
  geom_point() +
  geom_smooth(color = "#9e42f5", method = "lm", se = F) +
  annotate("text", x = 5.2, y = 0.25, hjust = 0, label = paste0("rho = ", cor_coef)) +
  annotate("text", x= 5.2, y = 0.24, hjust = 0, label = substitute(paste(italic("p"), " = 0.022"))) +
  ggtitle(paste("CD4 T memory - CCL18")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), 
        legend.title = element_blank(), legend.position = "none",
        aspect.ratio = 1,
        axis.title = element_text(face = "bold")) +
  xlab("log2 expression") +
  ylab("proportion")

ggsave("DSP1_Spearman_TIL-CD4Tmem_TLS-CCL18.pdf", device = "pdf",height = 4, width = 4, path = figdir)

########

a <- cbind(prop = prop_mat[,"CD4 T memory"], 
           chemo = cytokine_mat[,"CXCL9"]) %>% 
  as.data.frame

cor_coef <- round(cor_mat["CXCL9", "CD4 T memory"], digits = 2)
cor_pval <- round(cor_test_mat["CXCL9", "CD4 T memory"], digits = 3) %>% print

ggplot(a, aes(x = chemo, y = prop)) +
  geom_point() +
  geom_smooth(color = "#9e42f5", method = "lm", se = F) +
  annotate("text", x = 4.9, y = 0.25, hjust = 0, label = paste0("rho = ", cor_coef)) +
  annotate("text", x= 4.9, y = 0.24, hjust = 0, label = substitute(paste(italic("p"), " = 0.003"))) +
  ggtitle(paste("CD4 T memory - CXCL9")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), 
        legend.title = element_blank(), legend.position = "none",
        aspect.ratio = 1,
        axis.title = element_text(face = "bold")) +
  xlab("log2 expression") +
  ylab("proportion")

ggsave("DSP1_Spearman_TIL-CD4Tmem_TLS-CXCL9.pdf", device = "pdf", height = 4, width = 4, path = figdir)

########

a <- cbind(prop = prop_mat[,"CD4 T naive"], 
           chemo = cytokine_mat[,"CXCL10"]) %>% 
  as.data.frame

cor_coef <- round(cor_mat["CXCL10", "CD4 T naive"], digits = 2)
cor_pval <- round(cor_test_mat["CXCL10", "CD4 T naive"], digits = 3) %>% print

ggplot(a, aes(x = chemo, y = prop)) +
  geom_point() +
  geom_smooth(color = "#9e42f5", method = "lm", se = F) +
  annotate("text", x = 4.8, y = 0.28, hjust = 0, label = paste0("rho = ", cor_coef)) +
  annotate("text", x= 4.8, y = 0.26, hjust = 0, label = substitute(paste(italic("p"), " = 0.002"))) +
  ggtitle(paste("CD4 T naive - CXCL10")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), 
        legend.title = element_blank(), legend.position = "none",
        aspect.ratio = 1,
        axis.title = element_text(face = "bold")) +
  xlab("log2 expression") +
  ylab("proportion")

ggsave("DSP1_Spearman_TIL-CD4Tnaive_TLS-CXCL10.pdf", device = "pdf", height = 4, width = 4, path = figdir)

########

a <- cbind(prop = prop_mat[,"CD8 T memory"], 
           chemo = cytokine_mat[,"CXCL10"]) %>% 
  as.data.frame

cor_coef <- round(cor_mat["CXCL10", "CD8 T memory"], digits = 2)
cor_pval <- round(cor_test_mat["CXCL10", "CD8 T memory"], digits = 3) %>% print

ggplot(a, aes(x = chemo, y = prop)) +
  geom_point() +
  geom_smooth(color = "#9e42f5", method = "lm", se = F) +
  annotate("text", x = 4.7, y = 0.5, hjust = 0, label = paste0("rho = ", cor_coef)) +
  annotate("text", x= 4.7, y = 0.48, hjust = 0, label = substitute(paste(italic("p"), " = 0.013"))) +
  ggtitle(paste("CD8 T memory - CXCL10")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), 
        legend.title = element_blank(), legend.position = "none",
        aspect.ratio = 1,
        axis.title = element_text(face = "bold")) +
  xlab("log2 expression") +
  ylab("proportion")

ggsave("DSP1_Spearman_TIL-CD8Tmem_TLS-CXCL10.pdf", device = "pdf", height = 4, width = 4, path = figdir)

########

a <- cbind(prop = prop_mat[,"CD8 T memory"], 
           chemo = cytokine_mat[,"CXCL13"]) %>% 
  as.data.frame

cor_coef <- round(cor_mat["CXCL13", "CD8 T memory"], digits = 2)
cor_pval <- round(cor_test_mat["CXCL13", "CD8 T memory"], digits = 3) %>% print

ggplot(a, aes(x = chemo, y = prop)) +
  geom_point() +
  geom_smooth(color = "#9e42f5", method = "lm", se = F) +
  annotate("text", x = 5, y = 0.5, hjust = 0, label = paste0("rho = ", cor_coef)) +
  annotate("text", x= 5, y = 0.48, hjust = 0, label = substitute(paste(italic("p"), " = 0.025"))) +
  ggtitle(paste("CD8 T memory - CXCL13")) +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"), 
        legend.title = element_blank(), legend.position = "none",
        aspect.ratio = 1,
        axis.title = element_text(face = "bold")) +
  xlab("log2 expression") +
  ylab("proportion")

ggsave("DSP1_Spearman_TIL-CD8Tmem_TLS-CXCL13.pdf", device = "pdf", height = 4, width = 4, path = figdir)

```

# immune abundance mature vs immature

```{r}
order <- rownames(ann)[which(ann$segment == "CD45")]
test <- absolute[,order]
colnames(test) <- ann$name[which(ann$segment == "CD45")]
test<- t(test)
test %<>% 
  as.data.frame %>% 
  rownames_to_column("name") %>% 
  left_join(ann[which(ann$segment == "CD45"),], "name") %>% 
  select(name:segment, mature_pt2)

#cells in TIL, mat vs imat
test %>% 
  filter(region == "TIL") %>%
  na.omit %>% 
  pivot_longer(cols = `macrophages`:`Treg`, names_to = "cells", values_to = "abundance") %>%
  ggplot(aes(x = cells, y = abundance)) +
  geom_boxplot(aes(fill = mature_pt2)) +
  theme_classic () +
  ggtitle("TIL, mature vs immature") +
  theme_bp()

#cells in TLS, mat vs imat
test %>% 
  filter(region == "TLS") %>%
  na.omit %>% 
  pivot_longer(cols = `macrophages`:`Treg`, names_to = "cells", values_to = "abundance") %>%
  ggplot(aes(x = cells, y = abundance)) +
  geom_boxplot(aes(fill = mature_pt2)) +
  theme_classic () +
  ggtitle("TLS, mature vs immature") +
  theme_bp()

#cells in TIL, GC vs  PM
test %>% 
  filter(region == "TIL") %>%
  na.omit %>% 
  pivot_longer(cols = `macrophages`:`Treg`, names_to = "cells", values_to = "abundance") %>%
  ggplot(aes(x = cells, y = abundance)) +
  geom_boxplot(aes(fill = tissue)) +
  theme_classic () +
    ggtitle("TIL, GC vs  PM") +
  theme_bp()

#cells in TLS, GC vs  PM
test %>% 
  filter(region == "TLS") %>%
  na.omit %>% 
  pivot_longer(cols = `macrophages`:`Treg`, names_to = "cells", values_to = "abundance") %>%
  ggplot(aes(x = cells, y = abundance)) +
  geom_boxplot(aes(fill = tissue)) +
  theme_classic () +
   ggtitle("TLS, GC vs PM") +
  theme_bp()

```
# Boxplot subsets per TIL maturity state
```{r}
stat_test <- props_TIL_all_long %>%
  select(name:prop) %>%
  left_join(ann[which(ann$region == "TIL"),], by = "name") %>%
  select(name:region, mature_pt2) %>%
  filter(cells %in% coi) %>% 
  na.omit %>%
  mutate(mature_pt2 = factor(mature_pt2, 
                           levels = c("non-mature", "mature"))) %>%
  mutate(cells = factor(cells, 
                        levels = c("B naive", "B memory", "plasma cells", "CD4 T naive", 
                                   "CD4 T memory", "CD8 T naive", "CD8 T memory","Treg","macrophages", 
                                   "classical monocytes", "non-classical monocytes"))) %>%
  group_by(cells) %>%
  pairwise_wilcox_test(prop ~ mature_pt2, 
                       p.adjust.method = "fdr") %>%
  mutate(p.symbol = p.adj.signif) 

props_TIL_all_long %>%  
  select(name:prop) %>%
  left_join(ann[which(ann$region == "TIL"),], by = "name") %>%
  select(name:region, mature_pt2) %>%
  filter(cells %in% coi) %>% 
  na.omit %>%
  mutate(mature_pt2 = factor(mature_pt2, 
                           levels = c("non-mature", "mature"))) %>% 
  mutate(cells = factor(cells, 
                        levels = c("B naive", "B memory", "plasma cells", "CD4 T naive", 
                                   "CD4 T memory", "CD8 T naive", "CD8 T memory","Treg","macrophages", 
                                   "classical monocytes", "non-classical monocytes"))) %>% 
  ggplot(aes(x = cells, y = prop)) +
    geom_boxplot(aes(fill = mature_pt2),
                 position = position_dodge(0.8), 
                 width = 0.6, 
                 fatten = 1) +
    scale_fill_manual(values = ann_c_maturity2[[1]][1:4]) +
    stat_pvalue_manual(stat_test %>%
                         add_x_position(x = "cells") %>%
                         add_y_position(step.increase = 0.06),
                     hide.ns = T,
                     label = "p.symbol",
                     tip.length = 0.01, label.size = 6) +
    theme_classic() +
    theme_bp()+
    ylab("proportions") +
    scale_x_discrete(labels = label_wrap(10))+
    ylim(-0.05, 1)

ggsave("DSP1_boxplots TIL_mature pt.pdf", device = "pdf", height = 5, width = 11, path = figdir)

```


## LMM-DE analysis TLS vs TIL
```{r }
#calculate enrichment of panels
ssgsea_results <- GSVA::gsva(expr = assayDataElement(target_Data,
                                   elt = "log_q"),
                            gset.idx.list = immune_panels,
                            method = "ssgsea",
                            min.sz = 3,
                            max.sz = 500,
                            verbose = FALSE)

#subset target_data based on the genesets
geneSetObj <-
  NanoStringGeoMxSet(assayData = ssgsea_results,
                     phenoData = AnnotatedDataFrame(pData(target_Data)),
                     protocolData = protocolData(target_Data),
                     featureType = "GeneSet",
                     check = FALSE)

#make a new df with only the expression data of the genesetobject
pw_matrix<-assayDataElement(geneSetObj, elt = "exprs")

```

```{r LMM_of_pathway}
pData(geneSetObj)$dfRegion <- as.factor(pData(geneSetObj)$region) %in% c("TIL", "TLS")
pData(geneSetObj)$dfPatient <- factor(pData(geneSetObj)$patient,
                                       levels=c("P1","P2","P3","P4","P5","P6","P7","P8","P9","P10"))
pData(geneSetObj)$dfState <- as.factor(pData(geneSetObj)$region)
pData(geneSetObj)[["slide"]] <- factor(pData(geneSetObj)[["slide name"]])

lmm_ssgsea_results <- c()

#analysis_regions <- unique(pData(geneSetObj)$dfRegion)
analysis_regions <- TRUE

for(region in analysis_regions) {
  print(region)
  active_aois<-pData(geneSetObj)$dfRegion == region
  mixedOutmc <-
    mixedModelDE(geneSetObj[, active_aois],
                 elt = "exprs",
                 modelFormula = ~ dfState + (1 | dfPatient),
                 groupVar = "dfState",
                 multiCore = FALSE,
                 pAdjust = NULL
                 )
  
  # format results as data.frame
  #r_ssgsea_df <- do.call(rbind.data.frame, mixedOutmc["lsmeans", ])
  lsmeans <- lapply(mixedOutmc, function(x) {x$lsmeans})
  anova <- unlist(lapply(mixedOutmc, function(x) {x$anova}))
  b <- lapply(lsmeans, function(x) {x[1]})
  c <- lapply(lsmeans, function(x) {x[2]})
  r_ssgsea_df <- cbind.data.frame(unlist(b), unlist(c))
  names(r_ssgsea_df) <- c('Estimate', 'Pr(>|t|)')
 
  ssgsea_dfs <- rownames(r_ssgsea_df)
  r_ssgsea_df$Contrast <- anova
  #r_ssgsea_df$Genes <- immune_panels seems unreliable as gsva omits pathways if genes are not in data..
  
  # use lapply in case you have multiple levels of your df factor to
  # correctly associate gene name with it's row in the results table
  r_ssgsea_df$Pathway <- rownames(r_ssgsea_df)
  r_ssgsea_df$Subset <- region
  
  r_ssgsea_df$FDR <- p.adjust(r_ssgsea_df$`Pr(>|t|)`, method = "fdr")
  r_ssgsea_df <- r_ssgsea_df[, c("Pathway", "Subset", "Contrast", "Estimate",
                                     "Pr(>|t|)", "FDR")]
  lmm_ssgsea_results <- rbind(lmm_ssgsea_results, r_ssgsea_df)
  
}

lmm_ssgsea_results<-merge(lmm_ssgsea_results,pw_gene_df)

```


Heatmap of DE pathways
```{r pw_heatmap, fig.height=15, fig.width=30}
options(scipen = 999) #disable scientific notation

active_pw<-filter(lmm_ssgsea_results, FDR < 0.05 )$Pathway

# order <- c(rownames(ann)[which(ann$region == "TIL")],
#            rownames(ann)[which(ann$maturity == "LA")], 
#            rownames(ann)[which(ann$maturity == "early TLS")],
#            rownames(ann)[which(ann$maturity == "PF-like TLS")],
#            rownames(ann)[which(ann$maturity == "SF-like TLS")])

order <- ann %>% filter(segment == "CD45") %>% arrange(region, mature_pt2) %>% select(aois) %>% unlist %>% as.vector

ann_c_mature_pt2 <- list("mature_pt2" = c("non-mature" = "#CBB9F9", "mature" = "#9066FA", "ambiguous" = "white"))

DE_im_pathways <- pheatmap(pw_matrix[active_pw,order],
           scale = "row", gaps_col = 20,
           show_rownames = TRUE, show_colnames = F,
           annotation_names_row = F,
           fontsize_row = 11,
           clustering_method = "complete",
           cluster_cols = F,
           #clustering_distance_cols = "correlation",
           clustering_distance_rows = "correlation",
           cutree_cols = 2, cutree_rows = 2,
           breaks = seq(-3, 3, 0.05),
           color = colorRampPalette(c("aquamarine", "black", "coral"))(120),
           cellheight = 12, cellwidth = 12, 
           border_color = "grey",
           main ="LMM DE: TLS vs TIL pathway analysis (FDR < 0.05)",
          annotation_colors = c(ann_c_TLS_TIL, ann_c_pt, ann_c_tissue, ann_c_panels, ann_c_mature_pt2),
           annotation_col = pData(target_Data)[which( pData(target_Data)$segment == "CD45"), c("mature_pt2", "tissue","patient", "region")])


save_pheatmap_pdf(x = DE_im_pathways, file = paste0(figdir,"DSP1_DE_im_pathways_with maturityANN_hm_supervised.pdf")) 
```

## Tumor pathways (non-DE)
For Tumor region: signalling pathways

```{r warning = F}
tumor_signalling <-  subset(msigdbr_list, names(msigdbr_list) %in% categories$pathways[categories$panel == "Signaling Pathways"])

ssgsea_results_tumor <- GSVA::gsva(expr = assayDataElement(target_Data,
                                   elt = "log_q"),
                            gset.idx.list = tumor_signalling,
                            method = "ssgsea",
                            min.sz = 5,
                            max.sz = 500,
                            verbose = FALSE)
```

Heatmap
```{r tumor heatmap, fig.height=10, fig.width=15}
tumor_pathway_ph <- pheatmap(ssgsea_results_tumor[,active_aois_PanCK],
           scale = "row", 
           show_rownames = TRUE, show_colnames = FALSE,
           fontsize_row = 11,
           clustering_method = "complete",
           clustering_distance_cols = "correlation",
           clustering_distance_rows = "correlation",
           cutree_cols = 1, cutree_rows = 2,
           breaks = seq(-3, 3, 0.05),
           color = colorRampPalette(c("aquamarine", "black", "coral"))(120),
           annotation_colors = c(ann_c_tumor, ann_c_pt, ann_c_tissue),
           cellheight = 12, cellwidth = 12, 
           border_color = "grey",
           main = "Pathways of tumor cells (non-DE)",
           annotation_col = pData(target_Data)[, c("tissue","patient", "region")])
```

# Export annotations and data

```{r}
combined <- "DSP-TLS/Combined/"

saveRDS(target_Data, file = paste0(combined, "target_Data1.RDS"))
write_xlsx(ann, paste0(combined, "ann_1.xlsx"))
```


```{r}
knitr::knit_exit()

```

