---
title: "DSP-TLS cohort 1: Nanostring GeoMx QC"
author: "R. Franco Fernandez"
output: 
    html_document:
      code_download: true
      dev: "png"
      code_folding: hide
      toc: true
      toc_float:
        collapsed: false
        smooth_scroll: true
      toc_depth: 3
      wrap: sentence
editor_options: 
    markdown: 
    wrap: sentence
---

# 1. Aim

The aim of this script is to perform QC and generate a QC-ed object for further analyses. 
This pipeline is based on the standard workflow as referenced in the study.

# 2. General setup for processing GeoMx data

## 2.1. Loading libaries

```{r setup, include=FALSE, message=F, warning=F}
library(NanoStringNCTools)
library(GeomxTools)
library(GeoMxWorkflows)
library(SpatialDecon)
library(GSVA)
library(msigdbr) 
library(ggforce)
library(ggplot2)
library(cowplot)
library(umap)
library(Rtsne)
library(pheatmap)
library(ggrepel) 
library(scales)
library(DT)
library(plotly)
library(gridExtra)
library(RColorBrewer)
library(writexl)
library(ggpubr)
library(tidyverse)
library(reshape2)

knitr::opts_chunk$set(echo = TRUE)
```

## 2.2. loading base files

```{r loading_base_data}

datadir<-file.path("cohort_1/GEOMX_RNA_hsCTA_006")
```

**Take care you import a column with nuclei count sepperately if you want.**

```{r parse_files}
DCCFiles <- dir(file.path(datadir, "dccs"), pattern = ".dcc$",
                full.names = TRUE, recursive = TRUE)
PKCFiles <- unzip(zipfile = dir(file.path(datadir, "pkcs"), pattern = ".zip$",
                                full.names = TRUE, recursive = TRUE))

SampleAnnotationFile <-
  dir(file.path(datadir, "annotation"), pattern = "^[^~]",
      full.names = TRUE, recursive = TRUE)
```

## 2.3. load data

```{r load_data}
Data <-
  readNanoStringGeoMxSet(dccFiles = DCCFiles,
                         pkcFiles = PKCFiles,
                         phenoDataFile = SampleAnnotationFile,
                         phenoDataSheet = "Sheet1",
                         phenoDataDccColName = "Sample_ID",
                         protocolDataColNames = c("aoi", "roi"),
                         experimentDataColNames = c("panel"))
```

# 3. Study design and annotations

```{r annotate}
pkcs <- annotation(Data)
modules <- gsub(".pkc", "", pkcs)
knitr::kable(data.frame(PKCs = pkcs, modules = modules))
```

```{r}
paste("Reads from following runs used: ",unique(pData(protocolData(Data))$SeqSetId))

```

```{r select_annotations}
tissue <-'tissue'
names(pData(Data))[names(pData(Data)) == "segment...15"] <- "segment" #change faulty name
count_mat <- dplyr::count(pData(Data), tissue,patient,region,segment,`slide name`)

#define colors(
ann_colors = list(
  'slide name' = c("GC1_2"=brewer.pal(8,"Dark2")[1], "GC3_4"=brewer.pal(8,"Dark2")[2], "GC5_6"=brewer.pal(8,"Dark2")[3],"PM1_4"=brewer.pal(8,"Dark2")[4]),

  tissue = c("GC"=brewer.pal(12,"Set3")[1], "PM"=brewer.pal(12,"Set3")[2]),
  pheno = c("GC"=brewer.pal(12,"Set3")[1], "PM"=brewer.pal(12,"Set3")[2]),
  
  
  region = c("Tumor"=brewer.pal(6,"Set1")[1],"TIL"=brewer.pal(6,"Set1")[2],"TLS"=brewer.pal(6,"Set1")[3]),
 
  patient = c("P1"=brewer.pal(12,"Set3")[4],
            "P2"=brewer.pal(12,"Set3")[5],	
            "P3"=brewer.pal(12,"Set3")[6],
            "P4"=brewer.pal(12,"Set3")[7],
            "P5"=brewer.pal(12,"Set3")[8],
            "P6"=brewer.pal(12,"Set3")[9],
            "P7"=brewer.pal(12,"Set3")[10],
            "P8"=brewer.pal(12,"Set3")[11],
            "P9"=brewer.pal(12,"Set3")[12],
            "P10"=brewer.pal(12,"Paired")[1] 
            ),
  PNAd = c("yes" = "#43BD2E", "no" = "#BD2E2E"),
  
 `DC-LAMP` = c("yes" = "#BD2EAF", "no" = "#2EBDB6"),
 
 `germinal center`= c("yes" = "#BD2EAF", "no" = "#2EBDB6"), 
 
 FDC = c("yes" = "#43BD2E", "no" = "#BD2E2E"),
 
 BCL6 = c("yes" = "#43BD2E", "no" = "#BD2E2E")
     
)     

```

```{r sankey, fig.width=20,fig.height=11}
ann_names<- c("slide name", "segment", "tissue", "region", "patient")
ann_size<-length(ann_names)

count_mat <- pData(Data) %>% dplyr::count(pData(Data)[c(ann_names)])

test_gr <- gather_set_data(count_mat, 1:all_of(ann_size))
test_gr$x <- factor(test_gr$x, labels = ann_names)

ggplot(test_gr, height = 10, width = 10, aes(x, id = id, split = y, value = n)) +
  geom_parallel_sets(aes(fill = region), axis.width = 0.1) +
  geom_parallel_sets_axes(axis.width = 0.2) +
  geom_parallel_sets_labels(color = "white", size = 5) +
  theme_classic(base_size = 17) + 
  theme(legend.position = "bottom",
        axis.ticks.y = element_blank(),
        axis.line = element_blank(),
        axis.text.y = element_blank()) +
  scale_y_continuous(expand = expansion(0)) + 
  scale_x_discrete(expand = expansion(0)) +
  scale_fill_manual(values=ann_colors$region) +
  labs(x = "", y = "") +
  annotate(geom = "segment", x = 3.25, xend = 3.25, y = 10, 
           yend = 60, lwd = 2) 

saveRDS(test_gr, "test_gr_DSP1.RDS")
saveRDS(ann_colors, "color_list_DSP1.RDS")
```

# 4. QC & pre-processing

## 4.1. Set QC parameters

shift any expression counts with a value of 0 to 1 to enable in downstream transformations

```{r shift_counts}
Data <- shiftCountsOne(Data, useDALogic = TRUE)
```

Default QC cutoffs are commented in () adjacent to the respective parameters study-specific values were selected after visualizing the QC results in more detail below

```{r set_QC_params}
#set to match Geomx GUI software

QC_params <-
  list(minSegmentReads = 1000, # Minimum number of reads (1000)
       percentTrimmed = 80,    # Minimum % of reads trimmed (80%)
       percentStitched = 90,   # Minimum % of reads stitched (80%)
       percentAligned = 75,    # Minimum % of reads aligned (80%)
       percentSaturation = 50, # Minimum sequencing saturation (50%)
       minNegativeCount = 1,   # Minimum negative control counts (10)
       maxNTCCount = 9000,     # Maximum counts observed in NTC well (1000)
       minNuclei = 100,         # Minimum # of nuclei estimated (100)
       minArea = 1000)         # Minimum segment area (5000)

Data <-
  setSegmentQCFlags(Data, 
                    qcCutoffs = QC_params)        


```

Collate QC Results

```{r collate_QC_results}
QCResults <- protocolData(Data)[["QCFlags"]]
flag_columns <- colnames(QCResults)
QC_Summary <- data.frame(Pass = colSums(!QCResults[, flag_columns]),
                         Warning = colSums(QCResults[, flag_columns]))
QCResults$QCStatus <- apply(QCResults, 1L, function(x) {
  ifelse(sum(x) == 0L, "PASS", "WARNING")
})
QC_Summary["TOTAL FLAGS", ] <-
  c(sum(QCResults[, "QCStatus"] == "PASS"),
    sum(QCResults[, "QCStatus"] == "WARNING"))

col_by <- "patient"
```

## 4.2. Graphical summaries of QC statistics plot function

```{r QC_plotting}
QC_histogram <- function(assay_data = NULL,
                         annotation = NULL,
                         fill_by = NULL,
                         thr = NULL,
                         scale_trans = NULL) {
  plt <- ggplot(assay_data,
                aes_string(x = paste0("unlist(`", annotation, "`)"),
                           fill = fill_by)) +
    geom_histogram(bins = 50) +
    geom_vline(xintercept = thr, lty = "dashed", color = "black") +
    theme_bw() + guides(fill = "none") +
    facet_wrap(as.formula(paste("~", fill_by)), nrow = 4) +
    labs(x = annotation, y = "segments, #", title = annotation)
  if(!is.null(scale_trans)) {
    plt <- plt +
      scale_x_continuous(trans = scale_trans)
  }
  plt
}

QC_histogram(sData(Data), "Trimmed (%)", col_by, 80)
QC_histogram(sData(Data), "Stitched (%)", col_by, 80)
QC_histogram(sData(Data), "Aligned (%)", col_by, 75)

QC_histogram(sData(Data), "Saturated (%)", col_by, 50) +
  labs(title = "Sequencing Saturation (%)",
       x = "Sequencing Saturation (%)")

QC_histogram(sData(Data), "area", col_by, 1000, scale_trans = "log10")
QC_histogram(sData(Data), "nuclei", col_by, 100)

# calculate the negative geometric means for each module
negativeGeoMeans <- 
  esBy(negativeControlSubset(Data), 
       GROUP = "Module", 
       FUN = function(x) { 
         assayDataApply(x, MARGIN = 2, FUN = ngeoMean, elt = "exprs") 
       }) 
protocolData(Data)[["NegGeoMean"]] <- negativeGeoMeans
```

Explicitly copy the Negative geoMeans from sData to pData

```{r copy_geomeans}
negCols <- paste0("NegGeoMean_", modules)
pData(Data)[, negCols] <- sData(Data)[["NegGeoMean"]]
for(ann in negCols) {
  plt <- QC_histogram(pData(Data), ann, col_by, 2, scale_trans = "log10")
  print(plt)
}
```

Detatch neg_geomean columns ahead of aggregateCounts call

```{r}
pData(Data) <- pData(Data)[, !colnames(pData(Data)) %in% negCols]
```

Show all NTC values, Freq = \# of Segments with a given NTC count:

```{r}
knitr::kable(table(NTC_Count = sData(Data)$NTC),
      col.names = c("NTC Count", "# of Segments"))

knitr::kable(QC_Summary, caption = "QC Summary Table for each Segment")
```

Subsetting our dataset has removed samples which did not pass QC

```{r subsetting_QC_fails}
demoData <- Data[, QCResults$QCStatus == "PASS"]
dim(Data)
```

Generally keep the qcCutoffs parameters unchanged.
Set removeLocalOutliers to FALSE if you do not want to remove local outliers

```{r process_QC}
Data <- setBioProbeQCFlags(Data, 
                               qcCutoffs = list(minProbeRatio = 0.1,
                                                percentFailGrubbs = 20), 
                               removeLocalOutliers = FALSE)

ProbeQCResults <- fData(Data)[["QCFlags"]]
```

Define QC table for Probe QC

```{r define_qc_table}
qc_df <- data.frame(Passed = sum(rowSums(ProbeQCResults[, -1]) == 0),
                    Global = sum(ProbeQCResults$GlobalGrubbsOutlier),
                    Local = sum(rowSums(ProbeQCResults[, -2:-1]) > 0
                                & !ProbeQCResults$GlobalGrubbsOutlier))
```

Subset object to exclude all that did not pass Ratio & Global testing

```{r subset}
ProbeQCPassed <- 
  subset(Data, 
         fData(Data)[["QCFlags"]][,c("LowProbeRatio")] == FALSE &
           fData(Data)[["QCFlags"]][,c("GlobalGrubbsOutlier")] == FALSE)
dim(ProbeQCPassed)

Data <- ProbeQCPassed 
```

Check how many unique targets the object has

```{r unique_check}
length(unique(featureData(Data)[["TargetName"]]))
```

Collapse to targets

```{r collaps_targets}
target_Data <- aggregateCounts(Data)
dim(target_Data)

exprs(target_Data)[1:5, 1:2]
```

Define LOQ SD threshold and minimum value

```{r set_LSQ}
cutoff <- 2
minLOQ <- 2
```

## 4.3. Limit of Quantification

Calculate LOQ per module tested

```{r calculate_LOQ}
LOQ <- data.frame(row.names = colnames(target_Data))
for(module in modules) {
  vars <- paste0(c("NegGeoMean_", "NegGeoSD_"),
                 module)
  if(all(vars[1:2] %in% colnames(pData(target_Data)))) {
    LOQ[, module] <-
      pmax(minLOQ,
           pData(target_Data)[, vars[1]] * 
             pData(target_Data)[, vars[2]] ^ cutoff)
  }
}
pData(target_Data)$LOQ <- LOQ
```

## 4.4 Filtering

```{r filtering}

LOQ_Mat <- c()
for(module in modules) {
  ind <- fData(target_Data)$Module == module
  Mat_i <- t(esApply(target_Data[ind, ], MARGIN = 1,
                     FUN = function(x) {
                       x > LOQ[, module]
                     }))
  LOQ_Mat <- rbind(LOQ_Mat, Mat_i)
}
# ensure ordering since this is stored outside of the geomxSet
LOQ_Mat <- LOQ_Mat[fData(target_Data)$TargetName, ]
```

Save detection rate information to pheno data

```{r save_detectino_rate}
pData(target_Data)$GenesDetected <- 
  colSums(LOQ_Mat, na.rm = TRUE)
pData(target_Data)$GeneDetectionRate <-
  pData(target_Data)$GenesDetected / nrow(target_Data)
```

Determine detection thresholds: 1%, 5%, 10%, 15%, \>15%

```{r determine+thresholds}
pData(target_Data)$DetectionThreshold <- 
  cut(pData(target_Data)$GeneDetectionRate,
      breaks = c(0, 0.01, 0.05, 0.1, 0.15, 1),
      labels = c("<1%", "1-5%", "5-10%", "10-15%", ">15%"))

# stacked bar plot of different cut points (1%, 5%, 10%, 15%)
ggplot(pData(target_Data),
       aes(x = DetectionThreshold)) +
  geom_bar(aes(fill = patient)) +
  geom_text(stat = "count", aes(label = ..count..), vjust = -0.5) +
  theme_bw() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  labs(x = "Gene Detection Rate",
       y = "Segments, #",
       fill = "Segment Type")
```

cut percent genes detected at 1, 5, 10, 15

```{r cut_to_percent}
knitr::kable(table(pData(target_Data)$DetectionThreshold,
            pData(target_Data)$patient))

target_Data <-
  target_Data[, pData(target_Data)$GeneDetectionRate >= .1]

dim(target_Data)
```

**Select the annotations we want to show, use \`\` to surround column names with spaces or special symbols**

```{r select_annotations2}
count_mat <- dplyr::count(pData(Data), `slide name`, region, patient, tissue, segment)
ann <- pData(target_Data) %>% as.data.frame %>% dplyr::select(`slide name`, name:BCL6)
```

Simplify the slide names

```{r simplify_slide_names}
# gather the data and plot in order: class, slide name, region, segment
# test_gr <- gather_set_data(count_mat, 1:5)
# test_gr$x <-
#   factor(test_gr$x,
#          levels = c("slide name", "tissue","patient","region","segment"))
```

# 4.4.1 Gene Detection Rate

Calculate detection rate

```{r clc_detection_rate}
LOQ_Mat <- LOQ_Mat[, colnames(target_Data)]
fData(target_Data)$DetectedSegments <- rowSums(LOQ_Mat, na.rm = TRUE)
fData(target_Data)$DetectionRate <-
  fData(target_Data)$DetectedSegments / nrow(pData(target_Data))
```

Gene of interest detection table

```{r gene_of_interest_table}
goi <- c("PDCD1", "CD274", "IFNG", "CD8A", "CD68", "EPCAM",
         "KRT18", "NPHS1", "NPHS2", "CALB1", "CLDN8")
goi_df <- data.frame(
  Gene = goi,
  Number = fData(target_Data)[goi, "DetectedSegments"],
  DetectionRate = percent(fData(target_Data)[goi, "DetectionRate"]))
```

# 4.4.2 Gene Filtering

Plot detection rate

```{r plot_det_rate}
plot_detect <- data.frame(Freq = c(1, 5, 10, 20, 30, 50))
plot_detect$Number <-
  unlist(lapply(c(0.01, 0.05, 0.1, 0.2, 0.3, 0.5),
                function(x) {sum(fData(target_Data)$DetectionRate >= x)}))
plot_detect$Rate <- plot_detect$Number / nrow(fData(target_Data))
rownames(plot_detect) <- plot_detect$Freq

ggplot(plot_detect, aes(x = as.factor(Freq), y = Rate, fill = Rate)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = formatC(Number, format = "d", big.mark = ",")),
            vjust = 1.6, color = "black", size = 4) +
  scale_fill_gradient2(low = "orange2", mid = "lightblue",
                       high = "dodgerblue3", midpoint = 0.65,
                       limits = c(0,1),
                       labels = scales::percent) +
  theme_bw() +
  scale_y_continuous(labels = scales::percent, limits = c(0,1),
                     expand = expansion(mult = c(0, 0))) +
  labs(x = "% of Segments",
       y = "Genes Detected, % of Panel > LOQ")
```

Subset to target genes detected in at least 10% of the samples.
Also manually include the negative control probe, for downstream use

```{r subset_to_10p_detected_genes}
negativeProbefData <- subset(fData(target_Data), CodeClass == "Negative")
neg_probes <- unique(negativeProbefData$TargetName)
target_Data <- 
  target_Data[fData(target_Data)$DetectionRate >= 0.1 |
                    fData(target_Data)$TargetName %in% neg_probes, ]
dim(target_Data)

# retain only detected genes of interest
goi <- goi[goi %in% rownames(target_Data)]
```

# 5. Normalization

We will now normalize the GeoMx data for downstream visualizations and differential expression.
The two common methods for normalization of DSP-NGS RNA data are i) quartile 3 (Q3) or ii) background normalization.

Both of these normalization methods estimate a normalization factor per segment to bring the segment data distributions together.
More advanced methods for normalization and modeling are under active development.
However, for most studies, these methods are sufficient for understanding differences between biological classes of segments and samples.

Q3 normalization is typically the preferred normalization strategy for most DSP-NGS RNA studies.
Given the low negative probe counts in this particular dataset as shown during Segment QC, we would further avoid background normalization as it may be less stable.

Before normalization, we will explore the relationship between the upper quartile (Q3) of the counts in each segment with the geometric mean of the negative control probes in the data.
Ideally, there should be a separation between these two values to ensure we have stable measure of Q3 signal.
If you do not see sufficient separation between these values, you may consider more aggressive filtering of low signal segments/genes.

Graph Q3 value vs negGeoMean of Negatives

```{r lpot_q3_negGeoMean}
ann_of_interest <- "tissue"
Stat_data <- 
  data.frame(row.names = colnames(exprs(target_Data)),
             Segment = colnames(exprs(target_Data)),
             Annotation = pData(target_Data)[, ann_of_interest],
             Q3 = unlist(apply(exprs(target_Data), 2,
                               quantile, 0.75, na.rm = TRUE)),
             NegProbe = exprs(target_Data)[neg_probes, ])
Stat_data_m <- melt(Stat_data, measure.vars = c("Q3", "NegProbe"),
                    variable.name = "Statistic", value.name = "Value")

plt1 <- ggplot(Stat_data_m,
               aes(x = Value, fill = Statistic)) +
  geom_histogram(bins = 40) + theme_bw() +
  scale_x_continuous(trans = "log2") +
  facet_wrap(~Annotation, nrow = 1) + 
  scale_fill_brewer(palette = 3, type = "qual") +
  labs(x = "Counts", y = "Segments, #")

plt2 <- ggplot(Stat_data,
               aes(x = NegProbe, y = Q3, color = Annotation)) +
  geom_abline(intercept = 0, slope = 1, lty = "dashed", color = "darkgray") +
  geom_point() + guides(color = "none") + theme_bw() +
  scale_x_continuous(trans = "log2") + 
  scale_y_continuous(trans = "log2") +
  scale_fill_manual(values=ann_colors$tissue) +
  theme(aspect.ratio = 1) +
  labs(x = "Negative Probe GeoMean, Counts", y = "Q3 Value, Counts")

plt3 <- ggplot(Stat_data,
               aes(x = NegProbe, y = Q3 / NegProbe, color = Annotation)) +
  geom_hline(yintercept = 1, lty = "dashed", color = "darkgray") +
  geom_point() + theme_bw() +
  scale_x_continuous(trans = "log2") + 
  scale_y_continuous(trans = "log2") +
  scale_fill_manual(values=ann_colors$tissue) +
  theme(aspect.ratio = 1) +
  labs(x = "Negative Probe GeoMean, Counts", y = "Q3/NegProbe Value, Counts")

btm_row <- plot_grid(plt2, plt3, nrow = 1, labels = c("B", ""),
                     rel_widths = c(0.43,0.57))
plot_grid(plt1, btm_row, ncol = 1, labels = c("A", ""))

```

Q3 norm (75th percentile) for WTA/CTA with or without custom spike-ins

```{r q3_norm}
target_Data <- normalize(target_Data ,
                             norm_method = "quant", 
                             desiredQuantile = .75,
                             toElt = "q_norm")
#, data_type = "RNA" depricated after 4.1

```

Background normalization for WTA/CTA without custom spike-in

```{r bg_norm}
target_Data <- normalize(target_Data,
                             norm_method = "neg", 
                             fromElt = "exprs",
                             toElt = "neg_norm")

# , data_type = "RNA" depricated after 4.1
```

## 5.1 Visualize the first 10 segments with each normalization method {.tabset .tabset-pills}

raw counts

```{r,fig.width=20, fig.height=8}
boxplot(exprs(target_Data)[,1:91],
        col = "#9EDAE5", main = "Raw Counts",
        log = "y", names = 1:91, xlab = "Segment",
        ylab = "Counts, Raw")
```

Q3 normalized

```{r, fig.width=20, fig.height=8}
boxplot(assayDataElement(target_Data[,1:90], elt = "q_norm"),
        col = "#2CA02C", main = "Q3 Norm Counts",
        log = "y", names = 1:90, xlab = "Segment",
        ylab = "Counts, Q3 Normalized")
```

Negative probe normalization

```{r}
boxplot(assayDataElement(target_Data[,1:8], elt = "neg_norm"),
        col = "#FF7F0E", main = "Neg Norm Counts",
        log = "y", names = 1:8, xlab = "Segment",
        ylab = "Counts, Neg. Normalized")
```
# 6. Add log_q
```{r}
assayDataElement(object = target_Data, elt = "log_q") <-
  assayDataApply(target_Data, 2, FUN = log, base = 2, elt = "q_norm")
```


# 7. Save object
```{r}
saveRDS(target_Data, "cohort_1/Analysis/target_Data.RDS")

saveRDS(Data, "cohort_1/Analysis/Data.RDS")

sessionInfo()
```


```{r}
knitr::knit_exit()
```

